### 2018-05-17 paper/main.tex
Started reorganizing paper directory.

\documentclass[sigplan,10pt,review]{acmart}

%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[sigplan,10pt,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,10pt,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[sigplan,10pt]{acmart}\settopmatter{}

\settopmatter{printfolios=true,printccs=false,printacmref=false}

% TODO notes
\usepackage[draft]{todonotes}

%% Agda stuff
\usepackage[conor]{agda}
\newcommand{\AK}{\AgdaKeyword}
\newcommand{\AY}{\AgdaSymbol}
\newcommand{\AN}{\AgdaNumber}
\newcommand{\AS}{\AgdaSpace}
\newcommand{\AB}{\AgdaBound}
\newcommand{\AO}{\AgdaOperator}
\newcommand{\AI}{\AgdaInductiveConstructor}
\newcommand{\AC}{\AgdaCoinductiveConstructor}
\newcommand{\AD}{\AgdaDatatype}
\newcommand{\AF}{\AgdaFunction}
\newcommand{\AM}{\AgdaModule}
\newcommand{\AL}{\AgdaField}
\newcommand{\AR}{\AgdaArgument}
\newcommand{\AT}{\AgdaIndent}
\newcommand{\ARR}{\AgdaRecord}
\newcommand{\AP}{\AgdaPostulate}
\newcommand{\APT}{\AgdaPrimitiveType}

\newcommand{\nonterm}[1]{\hspace*{-0.1cm}\colorbox{orange!25}{#1}}
\newcommand{\hole}[1]{\colorbox{yellow!50}{\ensuremath{\bigbox_{#1}}}}
%% Conference information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
\acmYear{2017}
\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2017}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%\citestyle{acmauthoryear}  %% For author/year citations
%\citestyle{acmnumeric}     %% For numeric citations
%\setcitestyle{nosort}      %% With 'acmnumeric', to disable automatic
                            %% sorting of references within a single citation;
                            %% e.g., \cite{Smith99,Carpenter05,Baker12}
                            %% rendered as [14,5,2] rather than [2,5,14].
%\setcitesyle{nocompress}   %% With 'acmnumeric', to disable automatic
                            %% compression of sequential references within a
                            %% single citation;
                            %% e.g., \cite{Baker12,Baker14,Baker16}
                            %% rendered as [2,3,4] rather than [2-4].

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Note: Authors migrating a paper from traditional SIGPLAN
%% proceedings format to PACMPL format must update the
%% '\documentclass' and topmatter commands above; see
%% 'acmart-pacmpl-template.tex'.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption
%% ODER: format ==         = "\mathrel{==}"
%% ODER: format /=         = "\neq "
%
%
\makeatletter
\@ifundefined{lhs2tex.lhs2tex.sty.read}%
  {\@namedef{lhs2tex.lhs2tex.sty.read}{}%
   \newcommand\SkipToFmtEnd{}%
   \newcommand\EndFmtInput{}%
   \long\def\SkipToFmtEnd#1\EndFmtInput{}%
  }\SkipToFmtEnd

\newcommand\ReadOnlyOnce[1]{\@ifundefined{#1}{\@namedef{#1}{}}\SkipToFmtEnd}
\usepackage{amstext}
\usepackage{amssymb}
\usepackage{stmaryrd}
\DeclareFontFamily{OT1}{cmtex}{}
\DeclareFontShape{OT1}{cmtex}{m}{n}
  {<5><6><7><8>cmtex8
   <9>cmtex9
   <10><10.95><12><14.4><17.28><20.74><24.88>cmtex10}{}
\DeclareFontShape{OT1}{cmtex}{m}{it}
  {<-> ssub * cmtt/m/it}{}
\newcommand{\texfamily}{\fontfamily{cmtex}\selectfont}
\DeclareFontShape{OT1}{cmtt}{bx}{n}
  {<5><6><7><8>cmtt8
   <9>cmbtt9
   <10><10.95><12><14.4><17.28><20.74><24.88>cmbtt10}{}
\DeclareFontShape{OT1}{cmtex}{bx}{n}
  {<-> ssub * cmtt/bx/n}{}
\newcommand{\tex}[1]{\text{\texfamily#1}}	% NEU

\newcommand{\Sp}{\hskip.33334em\relax}

\newcommand{\Conid}[1]{\mathit{#1}}
\newcommand{\Varid}[1]{\mathit{#1}}
\newcommand{\anonymous}{\kern0.06em \vbox{\hrule\@width.5em}}
\newcommand{\plus}{\mathbin{+\!\!\!+}}
\newcommand{\bind}{\mathbin{>\!\!\!>\mkern-6.7mu=}}
\newcommand{\rbind}{\mathbin{=\mkern-6.7mu<\!\!\!<}}% suggested by Neil Mitchell
\newcommand{\sequ}{\mathbin{>\!\!\!>}}
\renewcommand{\leq}{\leqslant}
\renewcommand{\geq}{\geqslant}
\usepackage{polytable}

%mathindent has to be defined
\@ifundefined{mathindent}%
  {\newdimen\mathindent\mathindent\leftmargini}%
  {}%

\def\resethooks{%
  \global\let\SaveRestoreHook\empty
  \global\let\ColumnHook\empty}
\newcommand*{\savecolumns}[1][default]%
  {\g@addto@macro\SaveRestoreHook{\savecolumns[#1]}}
\newcommand*{\restorecolumns}[1][default]%
  {\g@addto@macro\SaveRestoreHook{\restorecolumns[#1]}}
\newcommand*{\aligncolumn}[2]%
  {\g@addto@macro\ColumnHook{\column{#1}{#2}}}

\resethooks

\newcommand{\onelinecommentchars}{\quad-{}- }
\newcommand{\commentbeginchars}{\enskip\{-}
\newcommand{\commentendchars}{-\}\enskip}

\newcommand{\visiblecomments}{%
  \let\onelinecomment=\onelinecommentchars
  \let\commentbegin=\commentbeginchars
  \let\commentend=\commentendchars}

\newcommand{\invisiblecomments}{%
  \let\onelinecomment=\empty
  \let\commentbegin=\empty
  \let\commentend=\empty}

\visiblecomments

\newlength{\blanklineskip}
\setlength{\blanklineskip}{0.66084ex}

\newcommand{\hsindent}[1]{\quad}% default is fixed indentation
\let\hspre\empty
\let\hspost\empty
\newcommand{\NB}{\textbf{NB}}
\newcommand{\Todo}[1]{$\langle$\textbf{To do:}~#1$\rangle$}

\EndFmtInput
\makeatother
%
%
%
%
%
%
% This package provides two environments suitable to take the place
% of hscode, called "plainhscode" and "arrayhscode". 
%
% The plain environment surrounds each code block by vertical space,
% and it uses \abovedisplayskip and \belowdisplayskip to get spacing
% similar to formulas. Note that if these dimensions are changed,
% the spacing around displayed math formulas changes as well.
% All code is indented using \leftskip.
%
% Changed 19.08.2004 to reflect changes in colorcode. Should work with
% CodeGroup.sty.
%
\ReadOnlyOnce{polycode.fmt}%
\makeatletter

\newcommand{\hsnewpar}[1]%
  {{\parskip=0pt\parindent=0pt\par\vskip #1\noindent}}

% can be used, for instance, to redefine the code size, by setting the
% command to \small or something alike
\newcommand{\hscodestyle}{}

% The command \sethscode can be used to switch the code formatting
% behaviour by mapping the hscode environment in the subst directive
% to a new LaTeX environment.

\newcommand{\sethscode}[1]%
  {\expandafter\let\expandafter\hscode\csname #1\endcsname
   \expandafter\let\expandafter\endhscode\csname end#1\endcsname}

% "compatibility" mode restores the non-polycode.fmt layout.

\newenvironment{compathscode}%
  {\par\noindent
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \let\hspre\(\let\hspost\)%
   \pboxed}%
  {\endpboxed\)%
   \par\noindent
   \ignorespacesafterend}

\newcommand{\compaths}{\sethscode{compathscode}}

% "plain" mode is the proposed default.
% It should now work with \centering.
% This required some changes. The old version
% is still available for reference as oldplainhscode.

\newenvironment{plainhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\hspre\(\let\hspost\)%
   \pboxed}%
  {\endpboxed%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

\newenvironment{oldplainhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \(\pboxed}%
  {\endpboxed\)%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

% Here, we make plainhscode the default environment.

\newcommand{\plainhs}{\sethscode{plainhscode}}
\newcommand{\oldplainhs}{\sethscode{oldplainhscode}}
\plainhs

% The arrayhscode is like plain, but makes use of polytable's
% parray environment which disallows page breaks in code blocks.

\newenvironment{arrayhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \(\parray}%
  {\endparray\)%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

\newcommand{\arrayhs}{\sethscode{arrayhscode}}

% The mathhscode environment also makes use of polytable's parray 
% environment. It is supposed to be used only inside math mode 
% (I used it to typeset the type rules in my thesis).

\newenvironment{mathhscode}%
  {\parray}{\endparray}

\newcommand{\mathhs}{\sethscode{mathhscode}}

% texths is similar to mathhs, but works in text mode.

\newenvironment{texthscode}%
  {\(\parray}{\endparray\)}

\newcommand{\texths}{\sethscode{texthscode}}

% The framed environment places code in a framed box.

\def\codeframewidth{\arrayrulewidth}
\RequirePackage{calc}

\newenvironment{framedhscode}%
  {\parskip=\abovedisplayskip\par\noindent
   \hscodestyle
   \arrayrulewidth=\codeframewidth
   \tabular{@{}|p{\linewidth-2\arraycolsep-2\arrayrulewidth-2pt}|@{}}%
   \hline\framedhslinecorrect\\{-1.5ex}%
   \let\endoflinesave=\\
   \let\\=\@normalcr
   \(\pboxed}%
  {\endpboxed\)%
   \framedhslinecorrect\endoflinesave{.5ex}\hline
   \endtabular
   \parskip=\belowdisplayskip\par\noindent
   \ignorespacesafterend}

\newcommand{\framedhslinecorrect}[2]%
  {#1[#2]}

\newcommand{\framedhs}{\sethscode{framedhscode}}

% The inlinehscode environment is an experimental environment
% that can be used to typeset displayed code inline.

\newenvironment{inlinehscode}%
  {\(\def\column##1##2{}%
   \let\>\undefined\let\<\undefined\let\\\undefined
   \newcommand\>[1][]{}\newcommand\<[1][]{}\newcommand\\[1][]{}%
   \def\fromto##1##2##3{##3}%
   \def\nextline{}}{\) }%

\newcommand{\inlinehs}{\sethscode{inlinehscode}}

% The joincode environment is a separate environment that
% can be used to surround and thereby connect multiple code
% blocks.

\newenvironment{joincode}%
  {\let\orighscode=\hscode
   \let\origendhscode=\endhscode
   \def\endhscode{\def\hscode{\endgroup\def\@currenvir{hscode}\\}\begingroup}
   %\let\SaveRestoreHook=\empty
   %\let\ColumnHook=\empty
   %\let\resethooks=\empty
   \orighscode\def\hscode{\endgroup\def\@currenvir{hscode}}}%
  {\origendhscode
   \global\let\hscode=\orighscode
   \global\let\endhscode=\origendhscode}%

\makeatother
\EndFmtInput
%

%% Agda keywords

%% Agda standard types

%% Constructors of the above types

%% Some standard functions
% format map    = "\AF{map}"       % Disabled

%% Non-colored stuff

%% Useful symbols

% fontsize of code snippets
\renewcommand\hscodestyle{%
   \setlength\leftskip{0.25cm}%
   \footnotesize
}

\begin{document}

\title{Verified and terminating dissections}

\author{Carlos Tom\'e Corti\~nas}
\affiliation{
  \department{Department of Information and Computing Sciences}  
  \institution{University of Utrecht}
  \country{The Netherlands}          
}
\email{first1.last1@inst1.edu}

\author{Wouter Swierstra}
\affiliation{
  \department{Department of Information and Computing Sciences}
  \institution{University of Utrecht}
  \country{The Netherlands}
}
\email{w.s.swierstra@uu.nl}

\begin{abstract}
The functional programming paradigm encourages a style of programming based on
  the use of higher-order functions over inductively defined datatypes. A fold is
  the prototypical example of such a function. However, its use for computation
  comes at a price.  Folds for branching datatypes, such as binary trees, are by
  definition not tail recursive functions.

  McBride has proposed a method called
  \emph{dissection}\cite{McBride:2008:CLM:1328438.1328474}, to
  transform a fold into its tail-recursive counterpart. Nevertheless,
  it is not clear why the resulting function terminates, nor it is
  clear that the transformation preserves the fold's semantics. In
  this paper we fill the gap by providing a fully machine checked
  proof of the construction using the proof assistant Agda.
\end{abstract}

\include{ccs.tex}

%% Keywords
%% comma separated list
\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission

\maketitle

% Bound variables

\section{Introduction}

The type of binary trees is one of the most simple, yet widespread used,
datastructures in functional programming.  Beyond its elegance and simplicity,
in its definition lies an embarasing truth: a fold over a binary tree is not a
tail recursive function. In order to understand the problem, let us introduce
the type of binary trees.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\AK{data}\;\AD{Expr}\;\mathbin{:}\;\AP{Set}\;\AK{where}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\AI{Val}\;{}\<[10]%
\>[10]{}\mathbin{:}\;{}\<[13]%
\>[13]{}\AD{\ensuremath{\mathbb{N}}}\;{}\<[19]%
\>[19]{}\to \;\AD{Expr}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\AI{Add}\;{}\<[10]%
\>[10]{}\mathbin{:}\;{}\<[13]%
\>[13]{}\AD{Expr}\;{}\<[19]%
\>[19]{}\to \;\AD{Expr}\;\to \;\AD{Expr}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

We can write an evaluation function that maps binary trees to natural numbers if
we were to interpret the constructor \ensuremath{\AI{Add}} as addition.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\AF{eval}\;\mathbin{:}\;\AD{Expr}\;\to \;\AD{\ensuremath{\mathbb{N}}}{}\<[E]%
\\
\>[3]{}\AF{eval}\;(\AI{Val}\;\AB{n})\;{}\<[21]%
\>[21]{}\mathrel{=}\;\AB{n}{}\<[E]%
\\
\>[3]{}\AF{eval}\;(\AI{Add}\;\AB{\ensuremath{e_1}}\;\AB{\ensuremath{e_2}})\;{}\<[21]%
\>[21]{}\mathrel{=}\;\AF{eval}\;\AB{\ensuremath{e_1}}\;\AF{+}\;\AF{eval}\;\AB{\ensuremath{e_2}}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The function \ensuremath{\AF{eval}} is compositional. The value of a node \ensuremath{\AI{Add}} is computed by
adding the values denoted by its subexpressions. And here lies the problem. The
operator \ensuremath{\AF{\_+\_}} needs both of its parameters to be evaluated before it can further
reduce. If the expression we want to compute over is very big this poses a
problem during runtime as the execution stack grows linearly with the size of
the input.

\todo[inline]{Maybe include something about Optimization through tco => get rid of
intermediate steps}

In order to solve the problem, we can make the execution stack explicit and
write a function that performs tail recursion over it. The idea is that we can
use a list to store both intemediate results and the subtress that still need to
be processed. We can define such a stack as follows:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{12}{@{}>{\hspre}l<{\hspost}@{}}%
\column{20}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\AK{data}\;\AD{Stack}\;\mathbin{:}\;\AP{Set}\;\AK{where}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\AI{Top}\;{}\<[12]%
\>[12]{}\mathbin{:}\;\AD{Stack}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\AI{Left}\;{}\<[12]%
\>[12]{}\mathbin{:}\;\AD{Expr}\;{}\<[20]%
\>[20]{}\to \;\AD{Stack}\;\to \;\AD{Stack}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\AI{Right}\;{}\<[12]%
\>[12]{}\mathbin{:}\;\AD{\ensuremath{\mathbb{N}}}\;{}\<[20]%
\>[20]{}\to \;\AD{Stack}\;\to \;\AD{Stack}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

Two mutually recursive functions that operate over the stack and tree are load
and unload. The former traverses a tree finding the leftmost leaf while the
latter dispathes over the stack by accumulating a partial result while looking
for the next subtree.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{23}{@{}>{\hspre}l<{\hspost}@{}}%
\column{32}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\AK{mutual}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\nonterm{\AF{load}}\;\mathbin{:}\;\AD{Expr}\;\to \;\AD{Stack}\;\to \;\AD{\ensuremath{\mathbb{N}}}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\AF{load}\;(\AI{Val}\;\AB{n})\;{}\<[23]%
\>[23]{}\AB{stk}\;\mathrel{=}\;\nonterm{\AF{unload}}\;\AB{n}\;\AB{stk}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\AF{load}\;(\AI{Add}\;\AB{\ensuremath{e_1}}\;\AB{\ensuremath{e_2}})\;{}\<[23]%
\>[23]{}\AB{stk}\;\mathrel{=}\;\nonterm{\AF{load}}\;\AB{\ensuremath{e_1}}\;(\AI{Left}\;\AB{\ensuremath{e_2}}\;\AB{stk}){}\<[E]%
\\[\blanklineskip]%
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\nonterm{\AF{unload}}\;\mathbin{:}\;\AD{\ensuremath{\mathbb{N}}}\;\to \;\AD{Stack}\;\to \;\AD{\ensuremath{\mathbb{N}}}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\AF{unload}\;\AB{v}\;{}\<[16]%
\>[16]{}\AI{Top}\;{}\<[32]%
\>[32]{}\mathrel{=}\;\AB{v}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\AF{unload}\;\AB{v}\;{}\<[16]%
\>[16]{}(\AI{Right}\;\AB{v'}\;\AB{stk})\;{}\<[32]%
\>[32]{}\mathrel{=}\;\nonterm{\AF{unload}}\;(\AB{v'}\;\AF{+}\;\AB{v})\;\AB{stk}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\AF{unload}\;\AB{v}\;{}\<[16]%
\>[16]{}(\AI{Left}\;\AB{r}\;\AB{stk})\;{}\<[32]%
\>[32]{}\mathrel{=}\;\nonterm{\AF{load}}\;\AB{r}\;(\AI{Right}\;\AB{v}\;\AB{stk}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

A tail recursive version of \ensuremath{\AF{eval}} can is defined by calling load with the empty
stack.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\AF{tail\text{-}rec\text{-}eval}\;\mathbin{:}\;\AD{Expr}\;\Varid{→}\;\AD{\ensuremath{\mathbb{N}}}{}\<[E]%
\\
\>[3]{}\AF{tail\text{-}rec\text{-}eval}\;\AB{e}\;\mathrel{=}\;\AF{load}\;\AB{e}\;\AI{Top}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The reader might have noticed that the names of unload and load are higlighted
with \nonterm{orange}. This is because Agda's termination checker flags the pair of
functions as possibly non-terminating. Indeed, the recursive calls are not
always performed over syntactically smaller arguments. If we assumed that it
terminates we still do not know if it is correct in the sense that for every
input \ensuremath{\AF{tail\text{-}rec\text{-}eval}} and \ensuremath{\AF{eval}} agree on the output.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\AF{correct}\;\mathbin{:}\;\forall\;(\AB{e}\;\mathbin{:}\;\AD{Expr})\;\to \;\AF{eval}\;\AB{e}\;\AD{\ensuremath{\equiv}}\;\AF{tail\text{-}rec\text{-}eval}\;\AB{e}{}\<[E]%
\\
\>[3]{}\AF{correct}\;\mathrel{=}\;\hole{0}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The main contribution of this paper is to show that we can both prove
termination and correctness of the construction in Agda. Specifically we show
how we can encode an ordering relation over stacks, show that the relation is
well founded and use it to define a tail recursive fold by structural recursion.
We discuss what are the main challenges that we have to overcome to convinde
Agda's termination checker that indeed something `goes smaller`. Moreover, our
approach allows us to get correctness almost for free.

McBride's insight is that we can calculate the type of stacks from the generic
description of a datatype by using Leibniz rules for derivatives. In later
sections of the paper, we discuss how to generalize our results to regular tree
datatypes and by doing so we show that McBride's intuition that the construction
is correct was true.

\section{Basic idea}

The functions \ensuremath{\AF{load}} and \ensuremath{\AF{unload}} are marked as non terminating because they are
not defined by structural recursion over their arguments. In particular, the
stack passed as an argument to the recursive call of \ensuremath{\AF{load}} in the definition of
\ensuremath{\AF{unload}} is structurally equal in size as the input stack.

Intuitively, \ensuremath{\AF{load}} and \ensuremath{\AF{unload}} fold the tree by traversing it from its
leftmost leaf to its rightmost using the stack to store both partial results and
the remaining subtrees to fold them as neccesary. The problem arises because the
stack is simply typed and any information about how the subtrees kept in the
stack relate to each other and to the original tree is lost once a subtree is
inserted onto the stack.

However, it is clear that virtually every node (either leaf or not) from the
original tree is visited at most twice during the computation. First when the
function \ensuremath{\AF{load}} decomposes it looking for its leftmost leaf and a second time
when \ensuremath{\AF{unload}} is accumulating over the stack searching for another subtree to
continue. This process is depicted in figure 1.

\begin{figure}[h]
  \includegraphics[scale=0.25]{figure1}
\end{figure}

We can argue that because there are finitely many nodes on a tree, \ensuremath{\AF{load}} and
\ensuremath{\AF{unload}} neccesarily terminate. The question is now, How can we encode this
information in such a way that Agda understand that the fold terminates?

The idea is that \ensuremath{\AF{load}} and \ensuremath{\AF{unload}} should not fold  the full input tree in one
go, but instead they will perform one step of the computation at a time.
Morover, by defining them by structural recursion over their arguments now they
are classified as terminating by the termination checker.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{30}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\AF{load}\;\mathbin{:}\;\AD{Expr}\;\to \;\AD{Stack}\;\to \;\AD{\ensuremath{\mathbb{N}}}\;\AF{\ensuremath{\times}}\;\AD{Stack}{}\<[E]%
\\
\>[3]{}\AF{load}\;(\AI{Val}\;\AB{n})\;{}\<[21]%
\>[21]{}\AB{stk}\;\mathrel{=}\;(\AB{n}\;\Varid{,}\;\AB{stk}){}\<[E]%
\\
\>[3]{}\AF{load}\;(\AI{Add}\;\AB{\ensuremath{e_1}}\;\AB{\ensuremath{e_2}})\;{}\<[21]%
\>[21]{}\AB{stk}\;\mathrel{=}\;\AF{load}\;\AB{\ensuremath{e_1}}\;(\AI{Left}\;\AB{\ensuremath{e_2}}\;\AB{stk}){}\<[E]%
\\[\blanklineskip]%
\>[3]{}\AF{unload}\;\mathbin{:}\;\AD{\ensuremath{\mathbb{N}}}\;\to \;\AD{Stack}\;\to \;(\AD{\ensuremath{\mathbb{N}}}\;\AF{\ensuremath{\times}}\;\AD{Stack})\;\AD{\ensuremath{\uplus}}\;\AD{\ensuremath{\mathbb{N}}}{}\<[E]%
\\
\>[3]{}\AF{unload}\;\AB{v}\;{}\<[14]%
\>[14]{}\AI{Top}\;{}\<[30]%
\>[30]{}\mathrel{=}\;\AI{inj\ensuremath{_2}}\;\AB{v}{}\<[E]%
\\
\>[3]{}\AF{unload}\;\AB{v}\;{}\<[14]%
\>[14]{}(\AI{Right}\;\AB{v'}\;\AB{stk})\;{}\<[30]%
\>[30]{}\mathrel{=}\;\AF{unload}\;(\AB{v'}\;\AF{+}\;\AB{v})\;\AB{stk}{}\<[E]%
\\
\>[3]{}\AF{unload}\;\AB{v}\;{}\<[14]%
\>[14]{}(\AI{Left}\;\AB{r}\;\AB{stk})\;{}\<[30]%
\>[30]{}\mathrel{=}\;\AI{inj\ensuremath{_1}}\;(\AF{load}\;\AB{r}\;(\AI{Right}\;\AB{v}\;\AB{stk})){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

For example, if we take the same tree in figure 1, after \ensuremath{\AF{load}} finds the
initial leftmost leaf we can apply one step of the new \ensuremath{\AF{unload}} that will end up
in the next leaf to the right.

\begin{figure}[h]
  \includegraphics[scale=0.25]{figure2}
\end{figure}

A tail recursive fold corrensponds to repeatedly applying the function \ensuremath{\AF{unload}}
until we find a \ensuremath{\AI{inj\ensuremath{_2}}} whose value is the result of folding the tree.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\AF{tail\text{-}rec\text{-}eval}\;\mathbin{:}\;\AD{Expr}\;\to \;\AD{\ensuremath{\mathbb{N}}}{}\<[E]%
\\
\>[3]{}\AF{tail\text{-}rec\text{-}eval}\;\AB{e}\;\mathrel{=}\;\AF{rec}\;(\AF{load}\;\AB{e}\;\AI{Top}){}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\AK{where}{}\<[E]%
\\
\>[5]{}\hsindent{2}{}\<[7]%
\>[7]{}\nonterm{\AF{rec}}\;\mathbin{:}\;(\AD{\ensuremath{\mathbb{N}}}\;\AF{\ensuremath{\times}}\;\AD{Stack})\;\to \;\AD{\ensuremath{\mathbb{N}}}{}\<[E]%
\\
\>[5]{}\hsindent{2}{}\<[7]%
\>[7]{}\AF{rec}\;(\AB{n}\;\Varid{,}\;\AB{stk})\;\AK{with}\;\AF{unload}\;\AB{n}\;\AB{stk}{}\<[E]%
\\
\>[5]{}\hsindent{2}{}\<[7]%
\>[7]{}\Varid{...}\;\mid \;\AI{inj\ensuremath{_1}}\;\AB{z'}\;\mathrel{=}\;\nonterm{\AF{rec}}\;\AB{z'}{}\<[E]%
\\
\>[5]{}\hsindent{2}{}\<[7]%
\>[7]{}\Varid{...}\;\mid \;\AI{inj\ensuremath{_2}}\;\AB{r}\;\mathrel{=}\;\AB{r}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The function \ensuremath{\AF{tail\text{-}rec\text{-}eval}} still does not pass the termination checker, The
variable \ensuremath{\AB{z'}} is not structurally smaller than \ensuremath{(\AB{n}\;\Varid{,}\;\AB{stk})}. However, now we can
refine it by using well founded recursion to make it structurally recursive by
performing the recursion over the accessibility predicate instead of the pair
\ensuremath{\AD{\ensuremath{\mathbb{N}}}\;\AF{\ensuremath{\times}}\;\AD{Stack}}.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\AF{tail\text{-}rec\text{-}eval}\;\mathbin{:}\;\AD{Expr}\;\to \;\AD{\ensuremath{\mathbb{N}}}{}\<[E]%
\\
\>[3]{}\AF{tail\text{-}rec\text{-}eval}\;\AB{e}\;\mathrel{=}\;\AF{rec}\;(\AF{load}\;\AB{e}\;\AI{Top})\;\hole{1}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\AK{where}{}\<[E]%
\\
\>[5]{}\hsindent{2}{}\<[7]%
\>[7]{}\AF{rec}\;\mathbin{:}\;(\AB{z}\;\mathbin{:}\;\AD{\ensuremath{\mathbb{N}}}\;\AF{\ensuremath{\times}}\;\AD{Stack})\;\to \;\AD{Acc}\;\AD{\_<\_}\;\AB{z}\;\to \;\AD{\ensuremath{\mathbb{N}}}{}\<[E]%
\\
\>[5]{}\hsindent{2}{}\<[7]%
\>[7]{}\AF{rec}\;(\AB{n}\;\Varid{,}\;\AB{stk})\;(\AI{acc}\;\Varid{rs})\;\AK{with}\;\AF{unload}\;\AB{n}\;\AB{stk}{}\<[E]%
\\
\>[5]{}\hsindent{2}{}\<[7]%
\>[7]{}\Varid{...}\;\mid \;\AI{inj\ensuremath{_1}}\;\AB{z'}\;\mathrel{=}\;\AF{rec}\;\AB{z'}\;(\Varid{rs}\;\hole{2}){}\<[E]%
\\
\>[5]{}\hsindent{2}{}\<[7]%
\>[7]{}\Varid{...}\;\mid \;\AI{inj\ensuremath{_2}}\;\AB{r}\;\mathrel{=}\;\AB{r}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

For the function above to work, we need to find a suitable definition for the
relation \ensuremath{\AD{\_<\_}} over pairs of \ensuremath{\AD{\ensuremath{\mathbb{N}}}\;\AF{\ensuremath{\times}}\;\AD{Stack}}, prove that applying \ensuremath{\AF{unload}}
results in an smaller element by the relation and finally show that the relation
is \ensuremath{\AF{Well\text{-}founded}}, so the call to \ensuremath{\AF{rec}} can be made in the first place. Before
any of that, we need to revisit Huet's notion of \emph{Zipper} and show how it
relates to what we are trying to achieve.

\subsection{Zippers up, Zippers down}

Huet introduced \emph{Zippers} to allow a tree datastructure to be efficiently
updated in a purely functional way. The idea is that any location on a tree,
either an internal node or a leaf, can be represented by a path to the root and
the subtree that hangs downwards.

The pair \ensuremath{\AD{\ensuremath{\mathbb{N}}}\;\AF{\ensuremath{\times}}\;\AD{Stack}} used to compute the tail recursive fold is nothing more
that a restricted version of the \emph{Zipper} where the locations can only be
leaves of the tree.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\AD{Zipper}\;\mathbin{:}\;\AP{Set}{}\<[E]%
\\
\>[3]{}\AD{Zipper}\;\mathrel{=}\;\AD{\ensuremath{\mathbb{N}}}\;\AF{\ensuremath{\times}}\;\AD{Stack}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

From a \ensuremath{\AD{Zipper}} we have to be able to reconstruct the original \ensuremath{\AD{Expr}} which
will be neccesary later on for the proof that the relation is well founded. For
this matter, we have to enhance the type of stacks to store not only the partial
results but also the expressions that where consumed in order to produce them.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
\column{26}{@{}>{\hspre}l<{\hspost}@{}}%
\column{27}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\AD{Stack}\;\mathbin{:}\;\AP{Set}{}\<[E]%
\\
\>[3]{}\AD{Stack}\;\mathrel{=}\;\AD{List}\;(\AD{Expr}\;\AD{\ensuremath{\uplus}}\;{}\<[26]%
\>[26]{}(\AD{\ensuremath{\Sigma}}\;\AD{\ensuremath{\mathbb{N}}}\;\AK{\ensuremath{\lambda}}\;\AB{n}\;\to {}\<[E]%
\\
\>[26]{}\hsindent{1}{}\<[27]%
\>[27]{}\AD{\ensuremath{\Sigma}}\;\AD{Expr}\;\AK{\ensuremath{\lambda}}\;\AB{e}\;\to \;\AF{eval}\;\AB{e}\;\AD{\ensuremath{\equiv}}\;\AB{n})){}\<[E]%
\\
\>[3]{}\AK{pattern}\;\AI{Left}\;\AB{\ensuremath{t}}\;{}\<[25]%
\>[25]{}\mathrel{=}\;\AI{inj\ensuremath{_1}}\;\AB{\ensuremath{t}}{}\<[E]%
\\
\>[3]{}\AK{pattern}\;\AI{Right}\;\AB{n}\;\AB{\ensuremath{t}}\;\AB{\ensuremath{eq}}\;{}\<[25]%
\>[25]{}\mathrel{=}\;\AI{inj\ensuremath{_2}}\;(\AB{n}\;\Varid{,}\;\AB{\ensuremath{t}}\;\Varid{,}\;\AB{\ensuremath{eq}}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The original expression for which a \ensuremath{\AD{\ensuremath{\mathbb{N}}}\;\AF{\ensuremath{\times}}\;\AD{Stack}} represents a position can be
reconstructed by forgeting that some part has already been evaluated.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{27}{@{}>{\hspre}l<{\hspost}@{}}%
\column{36}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\AF{plug\ensuremath{\Uparrow}}\;\mathbin{:}\;\AD{Expr}\;\to \;\AD{Stack}\;\to \;\AD{Expr}{}\<[E]%
\\
\>[3]{}\AF{plug\ensuremath{\Uparrow}}\;\AB{e}\;\AI{[]}\;{}\<[36]%
\>[36]{}\mathrel{=}\;\AB{e}{}\<[E]%
\\
\>[3]{}\AF{plug\ensuremath{\Uparrow}}\;\AB{e}\;(\AI{Left}\;\AB{\ensuremath{t}}\;{}\<[27]%
\>[27]{}\hspace{.4em}\AI{:\hspace{-.1em}:}\hspace{.4em}\;\AB{stk})\;{}\<[36]%
\>[36]{}\mathrel{=}\;\AF{plug\ensuremath{\Uparrow}}\;(\AI{Add}\;\AB{e}\;\AB{\ensuremath{t}})\;\AB{stk}{}\<[E]%
\\
\>[3]{}\AF{plug\ensuremath{\Uparrow}}\;\AB{e}\;(\AI{Right}\;\AB{n}\;\AB{\ensuremath{t}}\;\AB{\ensuremath{eq}}\;{}\<[27]%
\>[27]{}\hspace{.4em}\AI{:\hspace{-.1em}:}\hspace{.4em}\;\AB{stk})\;{}\<[36]%
\>[36]{}\mathrel{=}\;\AF{plug\ensuremath{\Uparrow}}\;(\AI{Add}\;\AB{\ensuremath{t}}\;\AB{e})\;\AB{stk}{}\<[E]%
\\[\blanklineskip]%
\>[3]{}\AF{plugZ\ensuremath{\Uparrow}}\;\mathbin{:}\;(\AD{\ensuremath{\mathbb{N}}}\;\AF{\ensuremath{\times}}\;\AD{Stack})\;\to \;\AD{Expr}{}\<[E]%
\\
\>[3]{}\AF{plugZ\ensuremath{\Uparrow}}\;(\AB{n}\;\Varid{,}\;\AB{stk})\;\mathrel{=}\;\AF{plug\ensuremath{\Uparrow}}\;(\AI{Val}\;\AB{n})\;\AB{stk}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

Our goal is to impose an ordering relation over elements of \ensuremath{\AD{Zipper}} such that
for any input the \ensuremath{\AF{unload}} function delivers a \ensuremath{\AD{Zipper}} that is smaller by the
relation when it does not terminate with a value. Because the folding happens
from left to right, we want the relation to order the leaves of the tree
accordingly, so the leftmost leaf is the greatest element and the rightmost the
smallest. Using the example from before, we number the leaves as follows:

\begin{figure}[h]
  \includegraphics[scale=0.25, angle=90]{figure3}
\end{figure}

Using the \ensuremath{\AD{Stack}} as a path from the leaf to the root of the tree is difficult
if not impossible to encode a smaller than relation that does not relate any two
elements. Such relation has to be defined by induction on the \ensuremath{\AD{Stack}} part of the
\ensuremath{\AD{Zipper}}. But for any two given stacks a priori we cannot know how many layers
we have to peel in order to reach a case where one of them is obviously smaller
that the other.

We can approach the problem by understanding the \ensuremath{\AD{Stack}} not as a path from the
leaf up to the root but from the root down to the leaf. This change of
perspective is realised with a new plug function that does the opposite of
\ensuremath{\AF{plug\ensuremath{\Uparrow}}}.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{28}{@{}>{\hspre}l<{\hspost}@{}}%
\column{37}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\AF{plug\ensuremath{\Downarrow}}\;\mathbin{:}\;\AD{Expr}\;\to \;\AD{Stack}\;\to \;\AD{Expr}{}\<[E]%
\\
\>[3]{}\AF{plug\ensuremath{\Downarrow}}\;\AB{e}\;\AI{[]}\;{}\<[37]%
\>[37]{}\mathrel{=}\;\AB{e}{}\<[E]%
\\
\>[3]{}\AF{plug\ensuremath{\Downarrow}}\;\AB{e}\;(\AI{Left}\;\AB{\ensuremath{t}}\;{}\<[28]%
\>[28]{}\hspace{.4em}\AI{:\hspace{-.1em}:}\hspace{.4em}\;\AB{stk})\;{}\<[37]%
\>[37]{}\mathrel{=}\;\AI{Add}\;(\AF{plug\ensuremath{\Downarrow}}\;\AB{e}\;\AB{stk})\;\AB{\ensuremath{t}}{}\<[E]%
\\
\>[3]{}\AF{plug\ensuremath{\Downarrow}}\;\AB{e}\;(\AI{Right}\;\AB{n}\;\anonymous \;\anonymous \;{}\<[28]%
\>[28]{}\hspace{.4em}\AI{:\hspace{-.1em}:}\hspace{.4em}\;\AB{stk})\;{}\<[37]%
\>[37]{}\mathrel{=}\;\AI{Add}\;\AB{\ensuremath{t}}\;(\AF{plug\ensuremath{\Downarrow}}\;\AB{e}\;\AB{stk}){}\<[E]%
\\[\blanklineskip]%
\>[3]{}\AF{plugZ\ensuremath{\Downarrow}}\;\mathbin{:}\;(\AD{\ensuremath{\mathbb{N}}}\;\AF{\ensuremath{\times}}\;\AD{Stack})\;\to \;\AD{Expr}{}\<[E]%
\\
\>[3]{}\AF{plugZ\ensuremath{\Downarrow}}\;(\AB{n}\;\Varid{,}\;\AB{stk})\;\mathrel{=}\;\AF{plug\ensuremath{\Downarrow}}\;(\AI{Val}\;\AB{n})\;\AB{stk}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

It is clear that both views of the \ensuremath{\AD{Zipper}} are related. Indeed, to transport
from one to the other we only have to reverse the stack. We show the equvalence
with the following lemma\footnote{The other way around only requires to use
BLABLA of \ensuremath{\AF{reverse}}}:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{23}{@{}>{\hspre}l<{\hspost}@{}}%
\column{44}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\AF{plug\ensuremath{\Downarrow}}\AF{\text{-}to\text{-}}\AF{plug\ensuremath{\Uparrow}}\;{}\<[23]%
\>[23]{}\mathbin{:}\;\forall\;(\AB{e}\;\mathbin{:}\;\AD{Expr})\;(\AB{stk}\;\mathbin{:}\;\AD{Stack}){}\<[E]%
\\
\>[23]{}\Varid{→}\;\AF{plug\ensuremath{\Downarrow}}\;\AB{e}\;\AB{stk}\;\AD{\ensuremath{\equiv}}\;{}\<[44]%
\>[44]{}\AF{plug\ensuremath{\Uparrow}}\;\AB{e}\;(\AF{reverse}\;\AB{stk}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

Why do we need this equivalence? The bottom up view of a \ensuremath{\AD{Zipper}} is suitable
for defining the tail recursive fold, alas to prove termination we have to use
use the top down view to describe the relation we need.

\subsection{A relation on Zipper}

The relation over elements of \ensuremath{\AD{Zipper}} is defined by induction on the \ensuremath{\AD{Stack}}.
If we start in the root of the tree, we can navigate downwards both stacks in
parallel removing their common prefix. Once we find an \ensuremath{\AI{Add}} where they
disagree then whether the first \ensuremath{\AD{Zipper}} is located in the left or right
subtree fully determines if its bigger or smaller than the other \ensuremath{\AD{Zipper}}.
The following type accounts for this explanation:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
\column{23}{@{}>{\hspre}l<{\hspost}@{}}%
\column{44}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\AK{data}\;\AD{\_<\_}\;\mathbin{:}\;\AD{Zipper}\;\to \;\AD{Zipper}\;\to \;\AP{Set}\;\AK{where}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\AI{<\text{-}Right}\;{}\<[14]%
\>[14]{}\mathbin{:}\;(\AB{\ensuremath{t_1}}\;\Varid{,}\;\AB{\ensuremath{s_1}})\;\AD{<}\;(\AB{\ensuremath{t_2}}\;\Varid{,}\;\AB{\ensuremath{s_2}})\;{}\<[E]%
\\
\>[14]{}\to \;{}\<[18]%
\>[18]{}(\AB{\ensuremath{t_1}}\;\Varid{,}\;\AI{Right}\;\Varid{l}\;\AB{n}\;\AB{\ensuremath{eq}}\;\hspace{.4em}\AI{:\hspace{-.1em}:}\hspace{.4em}\;\AB{\ensuremath{s_1}})\;\AD{<}\;(\AB{\ensuremath{t_2}}\;\Varid{,}\;\AI{Right}\;\Varid{l}\;\AB{n}\;\AB{\ensuremath{eq}}\;\hspace{.4em}\AI{:\hspace{-.1em}:}\hspace{.4em}\;\AB{\ensuremath{s_2}}){}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\AI{<\text{-}Left}\;{}\<[14]%
\>[14]{}\mathbin{:}\;(\AB{\ensuremath{t_1}}\;\Varid{,}\;\AB{\ensuremath{s_1}})\;\AD{<}\;(\AB{\ensuremath{t_2}}\;\Varid{,}\;\AB{\ensuremath{s_2}})\;{}\<[E]%
\\
\>[14]{}\to \;{}\<[18]%
\>[18]{}(\AB{\ensuremath{t_1}}\;\Varid{,}\;\AI{Left}\;\AB{r}\;\hspace{.4em}\AI{:\hspace{-.1em}:}\hspace{.4em}\;\AB{\ensuremath{s_1}})\;{}\<[44]%
\>[44]{}\AD{<}\;(\AB{\ensuremath{t_2}}\;\Varid{,}\;\AI{Left}\;\AB{r}\;\hspace{.4em}\AI{:\hspace{-.1em}:}\hspace{.4em}\;\AB{\ensuremath{s_2}}){}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\AI{<\text{-}Right\text{-}Left}\;{}\<[19]%
\>[19]{}\mathbin{:}\;{}\<[23]%
\>[23]{}(\AB{\ensuremath{t_1}'}\;\AD{\ensuremath{\equiv}}\;\AF{plug\ensuremath{\Downarrow}}\;(\Conid{Tip}\;\AB{\ensuremath{t_2}})\;\AB{\ensuremath{s_2}})\;{}\<[E]%
\\
\>[19]{}\to \;{}\<[23]%
\>[23]{}(\AB{\ensuremath{t_2}'}\;\AD{\ensuremath{\equiv}}\;\AF{plug\ensuremath{\Downarrow}}\;(\Conid{Tip}\;\AB{\ensuremath{t_1}})\;\AB{\ensuremath{s_1}})\;{}\<[E]%
\\
\>[19]{}\to \;{}\<[23]%
\>[23]{}(\AB{\ensuremath{t_1}}\;\Varid{,}\;\AI{Right}\;\AB{n}\;\AB{\ensuremath{t_1}'}\;\AB{\ensuremath{eq}}\;\hspace{.4em}\AI{:\hspace{-.1em}:}\hspace{.4em}\;\AB{\ensuremath{s_1}})\;\AD{<}\;(\AB{\ensuremath{t_2}}\;\Varid{,}\;\AI{Left}\;\AB{\ensuremath{t_2}'}\;\hspace{.4em}\AI{:\hspace{-.1em}:}\hspace{.4em}\;\AB{\ensuremath{s_2}}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

Having the relation defined, we turn our focus to prove that it is well founded.
This is an important step towards filling the holes that were left open in the
function \ensuremath{\AF{tail\text{-}rec\text{-}eval}}.

A relation is well founded iff all the descending chains starting from an
arbitrary element are finite. In a theorem prover such as Agda, an alternative
definition of well foundedness is used which is based on an accesibility
predicate.

We can try to prove that the relation is well founded by using an auxiliary
function that allows us to pattern match on the smaller than proof. When doing
so, the inputs are refined to concrete constructors. Normally the proof either
makes use of recursion over the proof or over the input, but in the case of the
\ensuremath{\AI{<\text{-}Right\text{-}Left}} constructor we do not have either option, because the smaller
element is not structurally related to the bigger and the proof does not have
any recursive structure to use.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{<-WF}\;\mathbin{:}\;\AF{Well\text{-}founded}\;\AD{\_<\_}{}\<[E]%
\\
\>[B]{}\Varid{<-WF}\;\Varid{x}\;\mathrel{=}\;\AI{acc}\;(\Varid{aux}\;\Varid{x}){}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\AK{where}{}\<[E]%
\\
\>[5]{}\hsindent{2}{}\<[7]%
\>[7]{}\Varid{aux}\;\mathbin{:}\;\Varid{∀}\;(\Varid{x}\;\mathbin{:}\;\AD{Zipper})\;{}\<[E]%
\\
\>[7]{}\hsindent{4}{}\<[11]%
\>[11]{}\to \;\Varid{∀}\;(\Varid{y}\;\mathbin{:}\;\AD{Zipper})\;\to \;\Varid{y}\;\AD{<}\;\Varid{x}\;\to \;\AD{Acc}\;\AD{\_<\_}\;\Varid{y}{}\<[E]%
\\
\>[5]{}\hsindent{2}{}\<[7]%
\>[7]{}\Varid{aux}\;\AS{.}\hspace*{-0.1cm}\;(\AB{\ensuremath{t_2}}\;\Varid{,}\;\AI{Left}\;\AB{\ensuremath{t_1}'}\;\hspace{.4em}\AI{:\hspace{-.1em}:}\hspace{.4em}\;\AB{\ensuremath{s_2}})\;\AS{.}\hspace*{-0.1cm}\;(\AB{\ensuremath{t_1}}\;\Varid{,}\;\AI{Right}\;\AB{n}\;\AB{\ensuremath{t_2}'}\;\AB{\ensuremath{eq}}\;\hspace{.4em}\AI{:\hspace{-.1em}:}\hspace{.4em}\;\AB{\ensuremath{s_1}})\;(\AI{<\text{-}Right\text{-}Left}\;\Varid{eq1}\;\Varid{eq2})\;\mathrel{=}\;\{\mskip1.5mu \mathbin{!!}\mskip1.5mu\}{}\<[E]%
\\
\>[5]{}\hsindent{2}{}\<[7]%
\>[7]{}\Varid{aux}\;\Varid{...}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The proof fails because in \ensuremath{\Varid{aux}} both Zippers \ensuremath{\Varid{x}} and \ensuremath{\Varid{y}} might very well be
locations of leaves belonging to different trees as far we know. Thanks to the 
use of dependent types, the property that a Zipper represents a location inside a 
concrete tree can be made explicit at the type level.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\AK{data}\;\AD{Zipper\ensuremath{\Downarrow}}\;(\AB{e}\;\mathbin{:}\;\AD{Expr})\;\mathbin{:}\;\AP{Set}\;\AK{where}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{\char92 \char95 ,\char92 \char95 }\;\mathbin{:}\;(\AB{z}\;\mathbin{:}\;\AD{Zipper})\;\to \;\AF{plugZ\ensuremath{\Downarrow}}\;\AB{z}\;\AD{\ensuremath{\equiv}}\;\AB{e}\;\to \;\AD{Zipper\ensuremath{\Downarrow}}\;\AB{e}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

We write a relation that is enforced to only relate Zippers beloging to the same \ensuremath{\AD{Expr}} by
indexing the \ensuremath{\AD{Zipper\ensuremath{\Downarrow}}}

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\AK{data}\;\Conid{IxltOp}\;\mathbin{:}\;(\AB{e}\;\mathbin{:}\;\AD{Expr})\;\to \;\AD{Zipper\ensuremath{\Downarrow}}\;\AB{e}\;\to \;\AD{Zipper\ensuremath{\Downarrow}}\;\AB{e}\;\to \;\AP{Set}\;\AK{where}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{...}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The concrete details of the relation follow very much the one we gave before
with the exception that every case has attached a new piece of information
specifying the concrete \ensuremath{\AD{Expr}} to which both Zippers stand as locations.

The indexed relation is now suitable for proving well foundedness because by
introducing the equalities we are able to show how the structure of the two
input values are related so it is straightforward to use recursion. In concrete,
the case we explained before now can be proven by learning that \ensuremath{(\AB{\ensuremath{t_2}}\;\Varid{,}\;\AB{\ensuremath{s_2}})}
stands for a position on the left subtree, and \ensuremath{(\AB{\ensuremath{t_1}}\;\Varid{,}\;\AB{\ensuremath{s_1}})} on the right  of a
common\ensuremath{\AI{Add}} node.

% \begin{code}
%  <-WF : forall (e : Expr) Well-founded ltOp
%  <-WF x = acc (aux x)
%     where
%       aux : ∀ (x : Zipper)
%           -> ∀ (y : Zipper) -> y < x -> Acc ltOp y
%       aux dotted(t2 , Left t1' :: s2) dotted(t1 , Right n t2' eq :: s1) (<-Right-Left eq1 eq2) = {!!}
%       aux ...
% \end{code}

\todo[inline]{STOP HERE}

\subsection{Termination}
\subsection{Correctness}
\section{Regular universe}
  + Universe interpretation generic programming
  + Fixpoint
  + Example??
\section{Dissection}
  + Dissection in agda
  + Plug
  + Zipper up Zipper down
  + Make clear the separation between recursion in the functor level and
  the fix level
  + relation on dissection?

\section{Conclusion and future work}

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
  This material is based upon work supported by the
  \grantsponsor{GS100000001}{National Science
    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
  conclusions or recommendations expressed in this material are those
  of the author and do not necessarily reflect the views of the
  National Science Foundation.
\end{acks}

%% Bibliography
\bibliography{main}

%% Appendix
\appendix
\section{Appendix}

Text of appendix \ldots

\end{document}

### 2018-05-18 paper/main.tex
Changed todo formatting and fill-width.

@@
 % TODO notes
-\usepackage[draft]{todonotes}
+%\usepackage[draft]{todonotes}
+\usepackage{xcolor}
+\newcommand\todo[1]{\textcolor{red}{\textbf{TODO:~}#1}}
 
@@
 \begin{abstract}
-The functional programming paradigm encourages a style of programming based on
-  the use of higher-order functions over inductively defined datatypes. A fold is
-  the prototypical example of such a function. However, its use for computation
-  comes at a price.  Folds for branching datatypes, such as binary trees, are by
-  definition not tail recursive functions.
-
-  McBride has proposed a method called
-  \emph{dissection}\cite{McBride:2008:CLM:1328438.1328474}, to
-  transform a fold into its tail-recursive counterpart. Nevertheless,
-  it is not clear why the resulting function terminates, nor it is
-  clear that the transformation preserves the fold's semantics. In
-  this paper we fill the gap by providing a fully machine checked
-  proof of the construction using the proof assistant Agda.
+% The functional programming paradigm encourages a style of programming based on
+%   the use of higher-order functions over inductively defined datatypes. A fold is
+%   the prototypical example of such a function. However, its use for computation
+%   comes at a price.  Folds for branching datatypes, such as binary trees, are by
+%   definition not tail recursive functions.
+
+%   McBride has proposed a method called
+%   \emph{dissection}\cite{McBride:2008:CLM:1328438.1328474}, to
+%   transform a fold into its tail-recursive counterpart. Nevertheless,
+%   it is not clear why the resulting function terminates, nor it is
+%   clear that the transformation preserves the fold's semantics. In
+%   this paper we fill the gap by providing a fully machine checked
+%   proof of the construction using the proof assistant Agda.
+  \todo{Write abstract}
 \end{abstract}
 
-\include{ccs.tex}
+\include{ccs}
 
@@
 %% comma separated list
-\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission
+\keywords{datatype generic programming, catamorphisms, dissection,
+  dependent types, Agda, \todo{Keywords?}}
 
@@
 locations of leaves belonging to different trees as far we know. Thanks to the
-use of dependent types, the property that a Zipper represents a location inside a 
+use of dependent types, the property that a Zipper represents a position inside a
 concrete tree can be made explicit at the type level.
@@
 
-We write a relation that is enforced to only relate Zippers beloging to the same \ensuremath{\AD{Expr}} by
-indexing the \ensuremath{\AD{Zipper\ensuremath{\Downarrow}}}
+We write a relation that is enforced to only relate Zippers beloging to the same
+\ensuremath{\AD{Expr}} by using a common value of that type as an of \ensuremath{\AD{Zipper\ensuremath{\Downarrow}}}
 
@@
 
-The concrete details of the relation follow very much the one we gave before
+The concrete details of the relation follow very much the one we gave before,
 with the exception that every case has attached a new piece of information
-specifying the concrete \ensuremath{\AD{Expr}} to which both Zippers stand as locations.
-
-The indexed relation is now suitable for proving well foundedness because by
-introducing the equalities we are able to show how the structure of the two
-input values are related so it is straightforward to use recursion. In concrete,
-the case we explained before now can be proven by learning that \ensuremath{(\AB{\ensuremath{t_2}}\;\Varid{,}\;\AB{\ensuremath{s_2}})}
-stands for a position on the left subtree, and \ensuremath{(\AB{\ensuremath{t_1}}\;\Varid{,}\;\AB{\ensuremath{s_1}})} on the right  of a
-common\ensuremath{\AI{Add}} node.
-
-% \begin{code}
-%  <-WF : forall (e : Expr) Well-founded ltOp
-%  <-WF x = acc (aux x)
-%     where
-%       aux : ∀ (x : Zipper)
-%           -> ∀ (y : Zipper) -> y < x -> Acc ltOp y
-%       aux dotted(t2 , Left t1' :: s2) dotted(t1 , Right n t2' eq :: s1) (<-Right-Left eq1 eq2) = {!!}
-%       aux ...
-% \end{code}
+specifying the concrete \ensuremath{\AD{Expr}} obtained by plugging both Zippers.
+
+The new version of the relation is suitable for proving well foundedness because
+we can pattern match on the equality included in the \ensuremath{\AD{Zipper\ensuremath{\Downarrow}}} type to show how 
+the overall structure decreases. This allows us to use the recursion we need to
+complete the proof.
+In particular, the case we were not able to prove before, now can be proven by learning that
+\ensuremath{(\AB{\ensuremath{t_2}}\;\Varid{,}\;\AB{\ensuremath{s_2}})} is a position on the left subtree while \ensuremath{(\AB{\ensuremath{t_1}}\;\Varid{,}\;\AB{\ensuremath{s_1}})} is on the
+right subtree of a common \ensuremath{\AI{Add}} node.
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{7}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\Varid{<-WF}\;\mathbin{:}\;\forall\;\AB{e}\;\to \;\AF{Well\text{-}founded}\;(\Conid{IxltOp}\;\AB{e}){}\<[E]%
+\\
+\>[3]{}\Varid{<-WF}\;\Varid{x}\;\mathrel{=}\;\AI{acc}\;(\Varid{aux}\;\AB{e}\;\Varid{x}){}\<[E]%
+\\
+\>[3]{}\hsindent{2}{}\<[5]%
+\>[5]{}\AK{where}{}\<[E]%
+\\
+\>[5]{}\hsindent{2}{}\<[7]%
+\>[7]{}\Varid{aux}\;\mathbin{:}\;\forall\;(\AB{e}\;\mathbin{:}\;\AD{Expr})\;\to \;\forall\;(\Varid{x}\;\mathbin{:}\;\AD{Zipper\ensuremath{\Downarrow}}\;\AB{e}){}\<[E]%
+\\
+\>[7]{}\hsindent{4}{}\<[11]%
+\>[11]{}\to \;\forall\;(\Varid{y}\;\mathbin{:}\;\AD{Zipper\ensuremath{\Downarrow}}\;\AB{e})\;\to \;\Varid{y}\;\AD{<}\;\Varid{x}\;\to \;\AD{Acc}\;(\Conid{IxltOp}\;\AB{e})\;\Varid{y}{}\<[E]%
+\\
+\>[5]{}\hsindent{2}{}\<[7]%
+\>[7]{}\Varid{aux}\;\AS{.}\hspace*{-0.1cm}\;(\AI{Add}\;(\Varid{plug}))\;\AS{.}\hspace*{-0.1cm}\;((\AB{\ensuremath{t_2}}\;\Varid{,}\;\AI{Left}\;\AB{\ensuremath{t_1}'}\;\hspace{.4em}\AI{:\hspace{-.1em}:}\hspace{.4em}\;\AB{\ensuremath{s_2}}))\;\Varid{,}\;\AI{refl})\;\AS{.}\hspace*{-0.1cm}\;((\AB{\ensuremath{t_1}}\;\Varid{,}\;\AI{Right}\;\AB{n}\;\AB{\ensuremath{t_2}'}\;\AB{\ensuremath{eq}}\;\hspace{.4em}\AI{:\hspace{-.1em}:}\hspace{.4em}\;\AB{\ensuremath{s_1}})\;\Varid{,}\;\Varid{eq2})\;{}\<[E]%
+\\
+\>[7]{}\hsindent{4}{}\<[11]%
+\>[11]{}(\AI{<\text{-}Right\text{-}Left}\;\Varid{eq1}\;\Varid{eq2})\;\mathrel{=}\;\{\mskip1.5mu \mathbin{!!}\mskip1.5mu\}\;{}\<[E]%
+\\
+\>[5]{}\hsindent{2}{}\<[7]%
+\>[7]{}\Varid{aux}\;\Varid{...}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+We have now the proof of well foundedness for the relation defined over top-down Zippers. We also have proven that
+there is an equivalence between top-down and bottom-up Zippers. We exploit it by using the top-down encoding for
+the termination proof while we use the bottom-up to actually compute in a tail recursive manner.
+
+Thus we prove a lemma stating that if we apply unload to a bottom-up Zipper and this results in another
+Zipper, then the result is smaller by the relation than the input. However, to show it we have to 
+convert them to the top down representation. In overall, what we have is the following lemma:
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[B]{}\Varid{unload-ltop}\;\mathbin{:}\;\forall\;\AB{n}\;\AB{\ensuremath{eq}}\;\Varid{s}\;\Varid{t'}\;\Varid{s'}\;\to \;\AF{unload}\;(\Conid{Tip}\;\AB{n})\;(\Conid{TipA}\;\AB{n})\;\AB{\ensuremath{eq}}\;\Varid{s}\;\AD{\ensuremath{\equiv}}\;\Varid{inj₁}\;(\Varid{t'}\;\Varid{,}\;\Varid{s'}){}\<[E]%
+\\
+\>[B]{}\hsindent{13}{}\<[13]%
+\>[13]{}\to \;(\Varid{t'}\;\Varid{,}\;\AF{reverse}\;\Varid{s'})\;\AD{\_<\_}\;(\AB{n}\;\Varid{,}\;\AF{reverse}\;\Varid{s}){}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+\subsection{Correctness}
+
+Indexing the \emph{Zipper} with an expression allow us to prove correcness of
+the transformation easily. The expression during the fold does not change, thus
+in every step of the computation the result of its evaluation remains constant.
+
+By using induction over the definition of unload, we can prove that when \ensuremath{\AF{unload}} 
+delivers a value, it corresponds to the result of evaluating of the input expression.
+In order to do so, we enrich the type of \ensuremath{\AF{unload}} to include the expression that has already
+been folded and we have its result.
+
+\begin{hscode}\SaveRestoreHook
+\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
+\>[3]{}\AF{unload}\;\mathbin{:}\;(\AB{e}\;\mathbin{:}\;\AD{Expr})\;\to \;(\AB{n}\;\mathbin{:}\;\AD{\ensuremath{\mathbb{N}}})\;\to \;\AF{eval}\;\AB{e}\;\AD{\ensuremath{\equiv}}\;\AB{n}\;\to \;\AD{Stack}\;\to \;(\AD{\ensuremath{\mathbb{N}}}\;\AF{\ensuremath{\times}}\;\AD{Stack})\;\AD{\ensuremath{\uplus}}\;\AD{\ensuremath{\mathbb{N}}}{}\<[E]%
+\\[\blanklineskip]%
+\>[3]{}\Varid{unload-correct}\;{}\<[19]%
+\>[19]{}\mathbin{:}\;\forall\;(\AB{e}\;\mathbin{:}\;\AD{Expr})\;(\AB{n}\;\mathbin{:}\;\AD{\ensuremath{\mathbb{N}}})\;(\AB{\ensuremath{eq}}\;\mathbin{:}\;\AF{eval}\;\AB{e}\;\AD{\ensuremath{\equiv}}\;\AB{n})\;(\Varid{s}\;\mathbin{:}\;\AD{Stack})\;(\Varid{x}\;\mathbin{:}\;\AD{\ensuremath{\mathbb{N}}}){}\<[E]%
+\\
+\>[19]{}\to \;\AF{unload}\;\AB{e}\;\AB{n}\;\AB{\ensuremath{eq}}\;\Varid{s}\;\Varid{≡}\;\AI{inj\ensuremath{_2}}\;\Varid{x}\;\to \;\AF{eval}\;\AB{e}\;\AD{\ensuremath{\equiv}}\;\Varid{x}{}\<[E]%
+\ColumnHook
+\end{hscode}\resethooks
+
+Proving correctness of the whole transformation amounts to show that it holds for the
+auxiliary recursor that we use to write the function \ensuremath{\AF{tail\text{-}rec\text{-}eval}}. We use well founded 
+recursion to do structural recursion over the accesibility predicate and use the lemma 
+\ensuremath{\Varid{unload-correct}} in the base case.
 
@@
 
-\subsection{Termination}
-\subsection{Correctness}
 \section{Regular universe}
@@
 \end{document}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: t
+%%% TeX-command-default: "lagda2pdf"
+%%% End:

### 2018-05-18 paper/main.tex
Edits to section 1.

@@
 %% use defaults for review submission.
-\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
-\acmYear{2017}
-\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
-\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
+\acmConference[TyDe'18]{ACM SIGPLAN Workshop on Type-driven
+Development (TyDe)}{January 01--03, 2017}{St. Louis, MO, USA}
+\acmYear{2018} \acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
+\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn} \startPage{1}
 
@@
 
-\title{Verified and terminating dissections}
+\title{Dissection: verified and terminating}
 
@@
 \begin{abstract}
-% The functional programming paradigm encourages a style of programming based on
-%   the use of higher-order functions over inductively defined datatypes. A fold is
-%   the prototypical example of such a function. However, its use for computation
-%   comes at a price.  Folds for branching datatypes, such as binary trees, are by
-%   definition not tail recursive functions.
-
-%   McBride has proposed a method called
-%   \emph{dissection}\cite{McBride:2008:CLM:1328438.1328474}, to
-%   transform a fold into its tail-recursive counterpart. Nevertheless,
-%   it is not clear why the resulting function terminates, nor it is
-%   clear that the transformation preserves the fold's semantics. In
-%   this paper we fill the gap by providing a fully machine checked
-%   proof of the construction using the proof assistant Agda.
+  Lorem ipsum sic dolor amet.  Lorem ipsum sic dolor amet.  Lorem
+  ipsum sic dolor amet.  Lorem ipsum sic dolor amet.  Lorem ipsum sic
+  dolor amet.  Lorem ipsum sic dolor amet.  Lorem ipsum sic dolor
+  amet.
   \todo{Write abstract}
@@
 
-%% Keywords
-%% comma separated list
 \keywords{datatype generic programming, catamorphisms, dissection,
-  dependent types, Agda, \todo{Keywords?}}
+  dependent types, Agda, well-founded recursion \todo{Keywords?}}
 
@@
 
-The type of binary trees is one of the most simple, yet widespread used,
-datastructures in functional programming.  Beyond its elegance and simplicity,
-in its definition lies an embarasing truth: a fold over a binary tree is not a
-tail recursive function. In order to understand the problem, let us introduce
-the type of binary trees.
+Folds, or \emph{catamorphisms}, are a pervasive
+programming pattern. A fold generalizes many simple traversals over an
+algebraic data type. Functions implemented by means of a fold are
+\emph{compositional} and structurally recursive. Consider, for
+instance, the following expression data type, written in the
+dependently typed programming language Agda\todo{citation}:
 
@@
 \end{hscode}\resethooks
-
-We can write an evaluation function that maps binary trees to natural numbers if
-we were to interpret the constructor \ensuremath{\AI{Add}} as addition.
-
+%
+We can write a write a simple evaluator, mapping expressions to
+natural numbers, as follows:
+%
 \begin{hscode}\SaveRestoreHook
@@
 \end{hscode}\resethooks
-
-The function \ensuremath{\AF{eval}} is compositional. The value of a node \ensuremath{\AI{Add}} is computed by
-adding the values denoted by its subexpressions. And here lies the problem. The
-operator \ensuremath{\AF{\_+\_}} needs both of its parameters to be evaluated before it can further
-reduce. If the expression we want to compute over is very big this poses a
-problem during runtime as the execution stack grows linearly with the size of
-the input.
-
-\todo[inline]{Maybe include something about Optimization through tco => get rid of
-intermediate steps}
-
-In order to solve the problem, we can make the execution stack explicit and
-write a function that performs tail recursion over it. The idea is that we can
-use a list to store both intemediate results and the subtress that still need to
-be processed. We can define such a stack as follows:
+%
+In the case for \ensuremath{\AI{Add}\;\AB{\ensuremath{e_1}}\;\AB{\ensuremath{e_2}}}, the \ensuremath{\AF{eval}} function makes two recursive
+calls and sums their results. Such a function can be implemented by a
+fold, passing the addition and identity functions as the argument
+algebra.
+
+Unfortunately, all in the garden is not rosy. The operator \ensuremath{\AF{\_+\_}}
+needs both of its parameters to be fully evaluated before it can
+reduce further. When evaluating large expressions, the size of the
+stack used during execution grows linearly with the size of the input,
+potentially leading to a stack overflow, causing the execution of the
+program to halt unexpectedly.
+
+To address this problem, we can rewrite the evaluator to be
+\emph{tail recursive}. Modern compilers typically map tail recursive
+functions to machine code that runs in constant memory. To write such
+a tail recursive function, we need to introduce an explicit stack
+storing both intemediate results and the subtrees that have not yet
+been evaluated. We can define such a stack as follows:
 
@@
 
-Two mutually recursive functions that operate over the stack and tree are load
-and unload. The former traverses a tree finding the leftmost leaf while the
-latter dispathes over the stack by accumulating a partial result while looking
-for the next subtree.
+We can define a tail-recursive evaluation function by means of a
+pair of mutually recursive functions \ensuremath{\AF{load}} and \ensuremath{\AF{unload}}. The \ensuremath{\AF{load}}
+function traverses the input expression, pushing subtrees on the
+stack; the \ensuremath{\AF{unload}} function unloads the stack, while accumulating a
+(partial) result.
 
@@
 
-A tail recursive version of \ensuremath{\AF{eval}} can is defined by calling load with the empty
-stack.
+We can now define a tail recursive version of \ensuremath{\AF{eval}} by
+calling \ensuremath{\AF{load}} with an initially empty stack:
 
@@
 
-The reader might have noticed that the names of unload and load are higlighted
-with \nonterm{orange}. This is because Agda's termination checker flags the pair of
-functions as possibly non-terminating. Indeed, the recursive calls are not
-always performed over syntactically smaller arguments. If we assumed that it
-terminates we still do not know if it is correct in the sense that for every
-input \ensuremath{\AF{tail\text{-}rec\text{-}eval}} and \ensuremath{\AF{eval}} agree on the output.
-
-\begin{hscode}\SaveRestoreHook
-\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
-\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[3]{}\AF{correct}\;\mathbin{:}\;\forall\;(\AB{e}\;\mathbin{:}\;\AD{Expr})\;\to \;\AF{eval}\;\AB{e}\;\AD{\ensuremath{\equiv}}\;\AF{tail\text{-}rec\text{-}eval}\;\AB{e}{}\<[E]%
-\\
-\>[3]{}\AF{correct}\;\mathrel{=}\;\hole{0}{}\<[E]%
-\ColumnHook
-\end{hscode}\resethooks
-
-The main contribution of this paper is to show that we can both prove
-termination and correctness of the construction in Agda. Specifically we show
-how we can encode an ordering relation over stacks, show that the relation is
-well founded and use it to define a tail recursive fold by structural recursion.
-We discuss what are the main challenges that we have to overcome to convinde
-Agda's termination checker that indeed something `goes smaller`. Moreover, our
-approach allows us to get correctness almost for free.
-
-McBride's insight is that we can calculate the type of stacks from the generic
-description of a datatype by using Leibniz rules for derivatives. In later
-sections of the paper, we discuss how to generalize our results to regular tree
-datatypes and by doing so we show that McBride's intuition that the construction
-is correct was true.
+Implementing this tail recursive evaluator has come at a price: Agda's
+termination checker flags the \ensuremath{\AF{load}} and \ensuremath{\AF{unload}} functions as
+potentially non-terminating by highlighting them
+\nonterm{orange}. Indeed, in the very last clause of the \ensuremath{\AF{unload}}
+function a recursive call is made to arguments that are not
+syntactically smaller. Furthermore, it is not clear at all that the
+tail recursive evaluator produces the same result as our original
+one. It is precisely these questions that this paper tackles by
+means of the following novel contributions:
+
+\begin{itemize}
+\item We give a verified proof of termination of \ensuremath{\AF{tail\text{-}rec\text{-}eval}} using
+  a carefully chosen \emph{well-founded relation}
+  (Section~\ref{sec:basics}). After redefining \ensuremath{\AF{tail\text{-}rec\text{-}eval}} using
+  this relation, we prove that it agrees with the \ensuremath{\AF{eval}} function on
+  all inputs.
+\item We generalize this construction, inspired by McBride's
+  generalization of Danvy's work on abstract machines. To do so, we
+  define a universe of algebraic data types and a generic fold
+  (Section~\ref{sec:universe}). Subsequently we show how to turn any
+  structurally recursive function defined using a fold into a tail
+  recursive variant (Section~\ref{sec:dissection}). 
+\item Finally, we sketch how the proofs of termination and semantics
+  preservation from our example are generalized to the generic fold
+  over arbitrary types in our universe (Section~\ref{correctness}).
+\end{itemize}
+
+All the constructions and proofs presented in this paper have been
+implemented in Agda\todo{citation}. The corresponding code is freely
+available online.\footnote{\todo{url}}
 
 \section{Basic idea}
-
-The functions \ensuremath{\AF{load}} and \ensuremath{\AF{unload}} are marked as non terminating because they are
-not defined by structural recursion over their arguments. In particular, the
-stack passed as an argument to the recursive call of \ensuremath{\AF{load}} in the definition of
-\ensuremath{\AF{unload}} is structurally equal in size as the input stack.
-
-Intuitively, \ensuremath{\AF{load}} and \ensuremath{\AF{unload}} fold the tree by traversing it from its
-leftmost leaf to its rightmost using the stack to store both partial results and
-the remaining subtrees to fold them as neccesary. The problem arises because the
-stack is simply typed and any information about how the subtrees kept in the
-stack relate to each other and to the original tree is lost once a subtree is
-inserted onto the stack.
-
-However, it is clear that virtually every node (either leaf or not) from the
-original tree is visited at most twice during the computation. First when the
-function \ensuremath{\AF{load}} decomposes it looking for its leftmost leaf and a second time
-when \ensuremath{\AF{unload}} is accumulating over the stack searching for another subtree to
-continue. This process is depicted in figure 1.
+\label{sec:basics}
+The functions \ensuremath{\AF{load}} and \ensuremath{\AF{unload}} are marked as non terminating
+because they are not defined by structural recursion over their
+arguments. In particular, the stack passed as an argument to the
+recursive call of \ensuremath{\AF{load}} in the definition of \ensuremath{\AF{unload}} is structurally
+equal in size as the input stack.
+
+Intuitively, \ensuremath{\AF{load}} and \ensuremath{\AF{unload}} fold the tree by traversing it from
+its leftmost leaf to its rightmost using the stack to store both
+partial results and the remaining subtrees to fold them as
+neccesary. The problem arises because the stack is simply typed and
+any information about how the subtrees kept in the stack relate to
+each other and to the original tree is lost once a subtree is inserted
+onto the stack.
+
+However, it is clear that virtually every node (either leaf or not)
+from the original tree is visited at most twice during the
+computation. First when the function \ensuremath{\AF{load}} decomposes it looking for
+its leftmost leaf and a second time when \ensuremath{\AF{unload}} is accumulating over
+the stack searching for another subtree to continue. This process is
+depicted in figure 1.
 
@@
 
-We can argue that because there are finitely many nodes on a tree, \ensuremath{\AF{load}} and
-\ensuremath{\AF{unload}} neccesarily terminate. The question is now, How can we encode this
-information in such a way that Agda understand that the fold terminates?
-
-The idea is that \ensuremath{\AF{load}} and \ensuremath{\AF{unload}} should not fold  the full input tree in one
-go, but instead they will perform one step of the computation at a time.
-Morover, by defining them by structural recursion over their arguments now they
-are classified as terminating by the termination checker.
+We can argue that because there are finitely many nodes on a tree,
+\ensuremath{\AF{load}} and \ensuremath{\AF{unload}} neccesarily terminate. The question is now, How
+can we encode this information in such a way that Agda understand that
+the fold terminates?
+
+The idea is that \ensuremath{\AF{load}} and \ensuremath{\AF{unload}} should not fold the full input
+tree in one go, but instead they will perform one step of the
+computation at a time.  Morover, by defining them by structural
+recursion over their arguments now they are classified as terminating
+by the termination checker.
 
@@
 
-For example, if we take the same tree in figure 1, after \ensuremath{\AF{load}} finds the
-initial leftmost leaf we can apply one step of the new \ensuremath{\AF{unload}} that will end up
-in the next leaf to the right.
+For example, if we take the same tree in figure 1, after \ensuremath{\AF{load}} finds
+the initial leftmost leaf we can apply one step of the new \ensuremath{\AF{unload}}
+that will end up in the next leaf to the right.
 
@@
 
-A tail recursive fold corrensponds to repeatedly applying the function \ensuremath{\AF{unload}}
-until we find a \ensuremath{\AI{inj\ensuremath{_2}}} whose value is the result of folding the tree.
+A tail recursive fold corrensponds to repeatedly applying the function
+\ensuremath{\AF{unload}} until we find a \ensuremath{\AI{inj\ensuremath{_2}}} whose value is the result of folding
+the tree.
 
@@
 
-The function \ensuremath{\AF{tail\text{-}rec\text{-}eval}} still does not pass the termination checker, The
-variable \ensuremath{\AB{z'}} is not structurally smaller than \ensuremath{(\AB{n}\;\Varid{,}\;\AB{stk})}. However, now we can
-refine it by using well founded recursion to make it structurally recursive by
-performing the recursion over the accessibility predicate instead of the pair
-\ensuremath{\AD{\ensuremath{\mathbb{N}}}\;\AF{\ensuremath{\times}}\;\AD{Stack}}.
+The function \ensuremath{\AF{tail\text{-}rec\text{-}eval}} still does not pass the termination
+checker, The variable \ensuremath{\AB{z'}} is not structurally smaller than \ensuremath{(\AB{n}\;\Varid{,}\;\AB{stk})}. However, now we can refine it by using well founded recursion
+to make it structurally recursive by performing the recursion over the
+accessibility predicate instead of the pair \ensuremath{\AD{\ensuremath{\mathbb{N}}}\;\AF{\ensuremath{\times}}\;\AD{Stack}}.
 
@@
 
-For the function above to work, we need to find a suitable definition for the
-relation \ensuremath{\AD{\_<\_}} over pairs of \ensuremath{\AD{\ensuremath{\mathbb{N}}}\;\AF{\ensuremath{\times}}\;\AD{Stack}}, prove that applying \ensuremath{\AF{unload}}
-results in an smaller element by the relation and finally show that the relation
-is \ensuremath{\AF{Well\text{-}founded}}, so the call to \ensuremath{\AF{rec}} can be made in the first place. Before
-any of that, we need to revisit Huet's notion of \emph{Zipper} and show how it
-relates to what we are trying to achieve.
+For the function above to work, we need to find a suitable definition
+for the relation \ensuremath{\AD{\_<\_}} over pairs of \ensuremath{\AD{\ensuremath{\mathbb{N}}}\;\AF{\ensuremath{\times}}\;\AD{Stack}}, prove that
+applying \ensuremath{\AF{unload}} results in an smaller element by the relation and
+finally show that the relation is \ensuremath{\AF{Well\text{-}founded}}, so the call to \ensuremath{\AF{rec}}
+can be made in the first place. Before any of that, we need to revisit
+Huet's notion of \emph{Zipper} and show how it relates to what we are
+trying to achieve.
 
@@
 
-Huet introduced \emph{Zippers} to allow a tree datastructure to be efficiently
-updated in a purely functional way. The idea is that any location on a tree,
-either an internal node or a leaf, can be represented by a path to the root and
-the subtree that hangs downwards.
-
-The pair \ensuremath{\AD{\ensuremath{\mathbb{N}}}\;\AF{\ensuremath{\times}}\;\AD{Stack}} used to compute the tail recursive fold is nothing more
-that a restricted version of the \emph{Zipper} where the locations can only be
-leaves of the tree.
+Huet introduced \emph{Zippers} to allow a tree datastructure to be
+efficiently updated in a purely functional way. The idea is that any
+location on a tree, either an internal node or a leaf, can be
+represented by a path to the root and the subtree that hangs
+downwards.
+
+The pair \ensuremath{\AD{\ensuremath{\mathbb{N}}}\;\AF{\ensuremath{\times}}\;\AD{Stack}} used to compute the tail recursive fold is
+nothing more that a restricted version of the \emph{Zipper} where the
+locations can only be leaves of the tree.
 
@@
 
-From a \ensuremath{\AD{Zipper}} we have to be able to reconstruct the original \ensuremath{\AD{Expr}} which
-will be neccesary later on for the proof that the relation is well founded. For
-this matter, we have to enhance the type of stacks to store not only the partial
-results but also the expressions that where consumed in order to produce them.
+From a \ensuremath{\AD{Zipper}} we have to be able to reconstruct the original \ensuremath{\AD{Expr}}
+which will be neccesary later on for the proof that the relation is
+well founded. For this matter, we have to enhance the type of stacks
+to store not only the partial results but also the expressions that
+where consumed in order to produce them.
 
@@
 
-The original expression for which a \ensuremath{\AD{\ensuremath{\mathbb{N}}}\;\AF{\ensuremath{\times}}\;\AD{Stack}} represents a position can be
-reconstructed by forgeting that some part has already been evaluated.
+The original expression for which a \ensuremath{\AD{\ensuremath{\mathbb{N}}}\;\AF{\ensuremath{\times}}\;\AD{Stack}} represents a
+position can be reconstructed by forgeting that some part has already
+been evaluated.
 
@@
 
-Our goal is to impose an ordering relation over elements of \ensuremath{\AD{Zipper}} such that
-for any input the \ensuremath{\AF{unload}} function delivers a \ensuremath{\AD{Zipper}} that is smaller by the
-relation when it does not terminate with a value. Because the folding happens
-from left to right, we want the relation to order the leaves of the tree
-accordingly, so the leftmost leaf is the greatest element and the rightmost the
-smallest. Using the example from before, we number the leaves as follows:
+Our goal is to impose an ordering relation over elements of \ensuremath{\AD{Zipper}}
+such that for any input the \ensuremath{\AF{unload}} function delivers a \ensuremath{\AD{Zipper}} that
+is smaller by the relation when it does not terminate with a
+value. Because the folding happens from left to right, we want the
+relation to order the leaves of the tree accordingly, so the leftmost
+leaf is the greatest element and the rightmost the smallest. Using the
+example from before, we number the leaves as follows:
 
@@
 
-Using the \ensuremath{\AD{Stack}} as a path from the leaf to the root of the tree is difficult
-if not impossible to encode a smaller than relation that does not relate any two
-elements. Such relation has to be defined by induction on the \ensuremath{\AD{Stack}} part of the
-\ensuremath{\AD{Zipper}}. But for any two given stacks a priori we cannot know how many layers
-we have to peel in order to reach a case where one of them is obviously smaller
-that the other.
-
-We can approach the problem by understanding the \ensuremath{\AD{Stack}} not as a path from the
-leaf up to the root but from the root down to the leaf. This change of
-perspective is realised with a new plug function that does the opposite of
-\ensuremath{\AF{plug\ensuremath{\Uparrow}}}.
+Using the \ensuremath{\AD{Stack}} as a path from the leaf to the root of the tree is
+difficult if not impossible to encode a smaller than relation that
+does not relate any two elements. Such relation has to be defined by
+induction on the \ensuremath{\AD{Stack}} part of the \ensuremath{\AD{Zipper}}. But for any two given
+stacks a priori we cannot know how many layers we have to peel in
+order to reach a case where one of them is obviously smaller that the
+other.
+
+We can approach the problem by understanding the \ensuremath{\AD{Stack}} not as a path
+from the leaf up to the root but from the root down to the leaf. This
+change of perspective is realised with a new plug function that does
+the opposite of \ensuremath{\AF{plug\ensuremath{\Uparrow}}}.
 
@@
 
-It is clear that both views of the \ensuremath{\AD{Zipper}} are related. Indeed, to transport
-from one to the other we only have to reverse the stack. We show the equvalence
-with the following lemma\footnote{The other way around only requires to use
-BLABLA of \ensuremath{\AF{reverse}}}:
+It is clear that both views of the \ensuremath{\AD{Zipper}} are related. Indeed, to
+transport from one to the other we only have to reverse the stack. We
+show the equvalence with the following lemma\footnote{The other way
+  around only requires to use BLABLA of \ensuremath{\AF{reverse}}}:
 
@@
 
-Why do we need this equivalence? The bottom up view of a \ensuremath{\AD{Zipper}} is suitable
-for defining the tail recursive fold, alas to prove termination we have to use
-use the top down view to describe the relation we need.
+Why do we need this equivalence? The bottom up view of a \ensuremath{\AD{Zipper}} is
+suitable for defining the tail recursive fold, alas to prove
+termination we have to use use the top down view to describe the
+relation we need.
 
@@
 
-The relation over elements of \ensuremath{\AD{Zipper}} is defined by induction on the \ensuremath{\AD{Stack}}.
-If we start in the root of the tree, we can navigate downwards both stacks in
-parallel removing their common prefix. Once we find an \ensuremath{\AI{Add}} where they
-disagree then whether the first \ensuremath{\AD{Zipper}} is located in the left or right
-subtree fully determines if its bigger or smaller than the other \ensuremath{\AD{Zipper}}.
-The following type accounts for this explanation:
+The relation over elements of \ensuremath{\AD{Zipper}} is defined by induction on the
+\ensuremath{\AD{Stack}}.  If we start in the root of the tree, we can navigate
+downwards both stacks in parallel removing their common prefix. Once
+we find an \ensuremath{\AI{Add}} where they disagree then whether the first \ensuremath{\AD{Zipper}}
+is located in the left or right subtree fully determines if its bigger
+or smaller than the other \ensuremath{\AD{Zipper}}.  The following type accounts for
+this explanation:
 
@@
 
-Having the relation defined, we turn our focus to prove that it is well founded.
-This is an important step towards filling the holes that were left open in the
-function \ensuremath{\AF{tail\text{-}rec\text{-}eval}}.
-
-A relation is well founded iff all the descending chains starting from an
-arbitrary element are finite. In a theorem prover such as Agda, an alternative
-definition of well foundedness is used which is based on an accesibility
-predicate.
-
-We can try to prove that the relation is well founded by using an auxiliary
-function that allows us to pattern match on the smaller than proof. When doing
-so, the inputs are refined to concrete constructors. Normally the proof either
-makes use of recursion over the proof or over the input, but in the case of the
-\ensuremath{\AI{<\text{-}Right\text{-}Left}} constructor we do not have either option, because the smaller
-element is not structurally related to the bigger and the proof does not have
+Having the relation defined, we turn our focus to prove that it is
+well founded.  This is an important step towards filling the holes
+that were left open in the function \ensuremath{\AF{tail\text{-}rec\text{-}eval}}.
+
+A relation is well founded iff all the descending chains starting from
+an arbitrary element are finite. In a theorem prover such as Agda, an
+alternative definition of well foundedness is used which is based on
+an accesibility predicate.
+
+We can try to prove that the relation is well founded by using an
+auxiliary function that allows us to pattern match on the smaller than
+proof. When doing so, the inputs are refined to concrete
+constructors. Normally the proof either makes use of recursion over
+the proof or over the input, but in the case of the \ensuremath{\AI{<\text{-}Right\text{-}Left}}
+constructor we do not have either option, because the smaller element
+is not structurally related to the bigger and the proof does not have
 any recursive structure to use.
@@
 
-The proof fails because in \ensuremath{\Varid{aux}} both Zippers \ensuremath{\Varid{x}} and \ensuremath{\Varid{y}} might very well be
-locations of leaves belonging to different trees as far we know. Thanks to the
-use of dependent types, the property that a Zipper represents a position inside a
-concrete tree can be made explicit at the type level.
+The proof fails because in \ensuremath{\Varid{aux}} both Zippers \ensuremath{\Varid{x}} and \ensuremath{\Varid{y}} might very
+well be locations of leaves belonging to different trees as far we
+know. Thanks to the use of dependent types, the property that a Zipper
+represents a position inside a concrete tree can be made explicit at
+the type level.
 
@@
 
-We write a relation that is enforced to only relate Zippers beloging to the same
-\ensuremath{\AD{Expr}} by using a common value of that type as an of \ensuremath{\AD{Zipper\ensuremath{\Downarrow}}}
+We write a relation that is enforced to only relate Zippers beloging
+to the same \ensuremath{\AD{Expr}} by using a common value of that type as an of
+\ensuremath{\AD{Zipper\ensuremath{\Downarrow}}}
 
@@
 
-The concrete details of the relation follow very much the one we gave before,
-with the exception that every case has attached a new piece of information
-specifying the concrete \ensuremath{\AD{Expr}} obtained by plugging both Zippers.
-
-The new version of the relation is suitable for proving well foundedness because
-we can pattern match on the equality included in the \ensuremath{\AD{Zipper\ensuremath{\Downarrow}}} type to show how 
-the overall structure decreases. This allows us to use the recursion we need to
-complete the proof.
-In particular, the case we were not able to prove before, now can be proven by learning that
-\ensuremath{(\AB{\ensuremath{t_2}}\;\Varid{,}\;\AB{\ensuremath{s_2}})} is a position on the left subtree while \ensuremath{(\AB{\ensuremath{t_1}}\;\Varid{,}\;\AB{\ensuremath{s_1}})} is on the
-right subtree of a common \ensuremath{\AI{Add}} node.
+The concrete details of the relation follow very much the one we gave
+before, with the exception that every case has attached a new piece of
+information specifying the concrete \ensuremath{\AD{Expr}} obtained by plugging both
+Zippers.
+
+The new version of the relation is suitable for proving well
+foundedness because we can pattern match on the equality included in
+the \ensuremath{\AD{Zipper\ensuremath{\Downarrow}}} type to show how the overall structure
+decreases. This allows us to use the recursion we need to complete the
+proof.  In particular, the case we were not able to prove before, now
+can be proven by learning that \ensuremath{(\AB{\ensuremath{t_2}}\;\Varid{,}\;\AB{\ensuremath{s_2}})} is a position on the left
+subtree while \ensuremath{(\AB{\ensuremath{t_1}}\;\Varid{,}\;\AB{\ensuremath{s_1}})} is on the right subtree of a common \ensuremath{\AI{Add}}
+node.
 
@@
 
-We have now the proof of well foundedness for the relation defined over top-down Zippers. We also have proven that
-there is an equivalence between top-down and bottom-up Zippers. We exploit it by using the top-down encoding for
-the termination proof while we use the bottom-up to actually compute in a tail recursive manner.
-
-Thus we prove a lemma stating that if we apply unload to a bottom-up Zipper and this results in another
-Zipper, then the result is smaller by the relation than the input. However, to show it we have to 
-convert them to the top down representation. In overall, what we have is the following lemma:
+We have now the proof of well foundedness for the relation defined
+over top-down Zippers. We also have proven that there is an
+equivalence between top-down and bottom-up Zippers. We exploit it by
+using the top-down encoding for the termination proof while we use the
+bottom-up to actually compute in a tail recursive manner.
+
+Thus we prove a lemma stating that if we apply unload to a bottom-up
+Zipper and this results in another Zipper, then the result is smaller
+by the relation than the input. However, to show it we have to convert
+them to the top down representation. In overall, what we have is the
+following lemma:
 
@@
 \subsection{Correctness}
-
-Indexing the \emph{Zipper} with an expression allow us to prove correcness of
-the transformation easily. The expression during the fold does not change, thus
-in every step of the computation the result of its evaluation remains constant.
-
-By using induction over the definition of unload, we can prove that when \ensuremath{\AF{unload}} 
-delivers a value, it corresponds to the result of evaluating of the input expression.
-In order to do so, we enrich the type of \ensuremath{\AF{unload}} to include the expression that has already
-been folded and we have its result.
+\label{sec:basic-correctness}
+Indexing the \emph{Zipper} with an expression allow us to prove
+correcness of the transformation easily. The expression during the
+fold does not change, thus in every step of the computation the result
+of its evaluation remains constant.
+
+By using induction over the definition of unload, we can prove that
+when \ensuremath{\AF{unload}} delivers a value, it corresponds to the result of
+evaluating of the input expression.  In order to do so, we enrich the
+type of \ensuremath{\AF{unload}} to include the expression that has already been
+folded and we have its result.
 
@@
 
-Proving correctness of the whole transformation amounts to show that it holds for the
-auxiliary recursor that we use to write the function \ensuremath{\AF{tail\text{-}rec\text{-}eval}}. We use well founded 
-recursion to do structural recursion over the accesibility predicate and use the lemma 
+Proving correctness of the whole transformation amounts to show that
+it holds for the auxiliary recursor that we use to write the function
+\ensuremath{\AF{tail\text{-}rec\text{-}eval}}. We use well founded recursion to do structural
+recursion over the accesibility predicate and use the lemma
 \ensuremath{\Varid{unload-correct}} in the base case.
@@
 \section{Regular universe}
+\label{sec:universe}
   + Universe interpretation generic programming
@@
 \section{Dissection}
+\label{sec:dissection}  
   + Dissection in agda
@@
   + relation on dissection?
+
+\section{Termination and correctness, generically}
+\label{sec:dissection}
 

### 2018-05-18 paper/main.tex
One last reread of intro.

@@
 
-Unfortunately, all in the garden is not rosy. The operator \ensuremath{\AF{\_+\_}}
+Unfortunately, not all in the garden is rosy. The operator \ensuremath{\AF{\_+\_}}
 needs both of its parameters to be fully evaluated before it can
@@
 
-To address this problem, we can rewrite the evaluator to be
+To address this problem, we can manually rewrite the evaluator to be
 \emph{tail recursive}. Modern compilers typically map tail recursive
@@
 We can define a tail-recursive evaluation function by means of a
-pair of mutually recursive functions \ensuremath{\AF{load}} and \ensuremath{\AF{unload}}. The \ensuremath{\AF{load}}
-function traverses the input expression, pushing subtrees on the
+pair of mutually recursive functions, \ensuremath{\AF{load}} and \ensuremath{\AF{unload}}. The \ensuremath{\AF{load}}
+function traverses the expressions, pushing subtrees on the
 stack; the \ensuremath{\AF{unload}} function unloads the stack, while accumulating a
@@
 tail recursive evaluator produces the same result as our original
-one. It is precisely these questions that this paper tackles by
-means of the following novel contributions:
+one. Writing such tail recursive functions by hand is both tedious and
+error-prone. It is precisely these issues that this paper tackles
+by making the following novel contributions:
 
@@
   all inputs.
-\item We generalize this construction, inspired by McBride's
-  generalization of Danvy's work on abstract machines. To do so, we
-  define a universe of algebraic data types and a generic fold
+\item We generalize this construction, inspired by McBride and Danvy's
+  work on calculating abstract machines from their recursive
+  counterparts. To do so, we define a universe of algebraic data types
+  and a generic fold operation
   (Section~\ref{sec:universe}). Subsequently we show how to turn any
-  structurally recursive function defined using a fold into a tail
-  recursive variant (Section~\ref{sec:dissection}). 
+  structurally recursive function defined using a fold into its tail
+  recursive counterpart (Section~\ref{sec:dissection}).
 \item Finally, we sketch how the proofs of termination and semantics
   preservation from our example are generalized to the generic fold
-  over arbitrary types in our universe (Section~\ref{correctness}).
+  over arbitrary types in our universe
+  (Section~\ref{correctness}).
+
 \end{itemize}
@@
 All the constructions and proofs presented in this paper have been
-implemented in Agda\todo{citation}. The corresponding code is freely
+implemented in and checked by Agda. The corresponding code is freely
 available online.\footnote{\todo{url}}

### 2018-05-20 paper/ccs.tex

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008</concept_id>
<concept_desc>Software and its engineering~General programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003456.10003457.10003521.10003525</concept_id>
<concept_desc>Social and professional topics~History of programming languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~General programming languages}
\ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

### 2018-05-20 Shake.hs
Track .fmt in Shake, split lhs2tex things into files, more content

@@ -15,6 +15,10 @@ lagda = "lagda"
 tex :: FilePath
 tex = "tex"
 
+-- fmt file extension
+fmt :: FilePath
+fmt = "fmt"
+
 -- lagda files used to build the thesis
 thesis_lagda_files :: [String]
 thesis_lagda_files = [ "main" , "tree" , "problem" , "background"]
@@ -22,6 +26,9 @@ thesis_lagda_files = [ "main" , "tree" , "problem" , "background"]
 -- lagda files used for the paper
 paper_lagda_files :: [String]
 paper_lagda_files = [ "main" ]
+
+paper_fmt_files :: [String]
+paper_fmt_files = [ "paper" , "intro" , "generic" ]
 
 main :: IO ()
 main = shakeArgs shakeOptions $ do
@@ -31,6 +38,10 @@ main = shakeArgs shakeOptions $ do
     putNormal "Cleaning files in thesis"
     cmd_ "latexmk -c -cd thesis/main.tex"
     removeFilesAfter "thesis" ["*.tex", "*.bbl"]
+
+    putNormal "Cleaning files in paper"
+    cmd_ "latexmk -c -cd paper/main.tex"
+    removeFilesAfter "paper" ["*.tex", "*.bbl"]
 
   phony "thesis" $ do
     need ["thesis/main" <.> pdf]
@@ -55,9 +66,9 @@ main = shakeArgs shakeOptions $ do
 
   "paper/main" <.> pdf %> \out -> do
     putNormal "Building paper"
-    let files = [ "paper" </> file <.> tex | file <- paper_lagda_files]
-        bib   = "paper/main.bib"
-        fmt   = "paper/paper.fmt"
-        sty   = "paper/agda.sty"
-    need (bib : fmt : sty : files)
+    let files      = [ "paper" </> file <.> tex | file <- paper_lagda_files]
+        bib        = "paper/main.bib"
+        fmtFiles   = [ "paper" </> file <.> fmt | file <- paper_fmt_files]
+        sty        = "paper/agda.sty"
+    need (bib : sty : (files ++ fmtFiles))
     cmd_ "latexmk -pdf -cd paper/main.tex"

### 2018-05-22 Shake.hs
More content

@@ -62,7 +62,7 @@ main = shakeArgs shakeOptions $ do
         fmt   = "thesis/thesis.fmt"
         sty   = "thesis/agda.sty"
     need (bib : fmt : sty : files)
-    cmd_ "latexmk -pdf -cd -xelatex thesis/main.tex"
+    cmd_ "latexmk -f -pdf -cd -xelatex thesis/main.tex"
 
   "paper/main" <.> pdf %> \out -> do
     putNormal "Building paper"
@@ -71,4 +71,4 @@ main = shakeArgs shakeOptions $ do
         fmtFiles   = [ "paper" </> file <.> fmt | file <- paper_fmt_files]
         sty        = "paper/agda.sty"
     need (bib : sty : (files ++ fmtFiles))
-    cmd_ "latexmk -pdf -cd paper/main.tex"
+    cmd_ "latexmk -f -pdf -cd paper/main.tex"

### 2018-06-01 paper/figures/figure3.tex
added figures (but the last one)

\begin{tikzpicture}
\tikzset{
    ncbar angle/.initial=90,
    ncbar/.style={
        to path=(\tikztostart)
        -- ($(\tikztostart)!#1!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztotarget)$)
        -- ($(\tikztotarget)!($(\tikztostart)!#1!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztotarget)$)!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztostart)$)
        -- (\tikztotarget)
    },
    ncbar/.default=0.5cm,
}

\tikzset{square left brace/.style={ncbar=0.15cm}}
\tikzset{square right brace/.style={ncbar=-0.15cm}}
  \tikzstyle{every node}=[font=\footnotesize]
  \tikzstyle{level 1}=[level distance=10mm, sibling distance=24mm]
  \tikzstyle{level 2}=[level distance=10mm, sibling distance=14mm]
  \tikzstyle{level 3}=[level distance=10mm, sibling distance=14mm]
  \tikzstyle{level 3}=[level distance=10mm, sibling distance=14mm]
  \tikzstyle{load}=[->,thick,color=blue]
  \tikzstyle{unload}=[->,shorten <=1pt,thick,color=red]
	% \draw[step=1cm, very thin, gray] (-6,-6) grid (6,6);
  \node (root) {\AI{Add}}
  child {node (node1) {\AI{Add}} child {node [label=below:\pgftextcircled{5}] (node2) {\AI{Val}\AS{}\AN{3}}} 
                                 child {node (node3) {\AI{Add}} 
                                   child {node [label=below:\pgftextcircled{4}] (node4) {\AI{Val}\AS{}\AN{7}}}
                                   child {node [label=below right:\pgftextcircled{3}] (node5) {\AI{Val}\AS{}\AN{1}}}}} 
  child {node (node6) {\AI{Add}} child {node [label=below:\pgftextcircled{2}] (node7) {\AI{Val}\AS{}\AN{2}}}
                                 child {node [label=below right:\pgftextcircled{1}] (node8) {\AI{Val}\AS{}\AN{0}}}};

  \draw[load] (root) to [ bend right=45] (node1);
  \draw[load] (node1) to [ bend right=45] (node2);
  \draw[unload] (node2) to [ bend left=15] (node3);
  \draw[load] (node3) to [ bend right=45] (node4);
  \draw[unload] (node4) to [ bend left=15] (node5);
  \draw[load] (node5) to [ loop below] (node5);
  \draw[unload] (node5) to [ bend right=45] (node3);
  \draw[unload] (node3) to [ bend right=45] (node1);
  \draw[unload] (node1) to [ bend left=15] (node6);
  \draw[load] (node6) to [ bend right=45] (node7);
  \draw[unload] (node7) to [ bend left=15] (node8);
  \draw[load] (node8) to [ loop below] (node8);
  \draw[unload] (node8) to [ bend right=45] (node6);
  \draw[unload] (node6) to [ bend right=45] (root);
  \draw[step=0.1cm,gray] ($(node2)-(0.2,0.2)$) grid ($(node1)-(0.05,-0.15)$);
  \draw[step=0.1cm,gray] (node2) grid (node4);
  \draw[step=0.1cm,gray] (node4) grid (node3);
  \draw[step=0.1cm,gray] (node2) grid ($(node1)!0.25!(node3)$);
  \draw[step=0.1cm,gray] (node2) grid ($(node1)!0.5!(node3)$);
  \draw[step=0.1cm,gray] (node2) grid ($(node1)!0.75!(node3)$);
  \draw[step=0.1cm,gray] (node4) grid ($(node5)!0.75!(node3)$);
  \draw[step=0.1cm,gray] ($(node4)-(0.2,0.5)$) grid ($(node5)!0.51!(node3)$);

  \draw [thick] (-1.5,-5) to [square left brace ] (-1.5,-4);
  \draw [thick] (3.5,-5) to [square right brace ] (3.5,-4);
	\node at (-0.75,-4.5) (right1) {\AI{Right}\AS{}\AN{7},};
	\node at (0.6,-4.5)    (right2) {\AI{Right}\AS{}\AN{3},};
  \node at  (1.75,-4.5) {\AI{Left}};
	\node at (-2.5, -4.5) {\AN{1}\quad,};
  \tikzstyle{every node}=[font=\tiny]
  \tikzstyle{level 1}=[level distance=5mm, sibling distance=7mm]
  \node at (2.75,-4.25) (left1) {\AI{Add}} 
          child {node {\AI{Val}\AS{}\AN{2}}}
          child {node {\AI{Val}\AS{}\AN{0}}};

    % \draw [red, thick] (1,0) to [square right brace] (1,4);
\end{tikzpicture}

### 2018-06-05 paper/ccs.tex
Generated CCS

@@
 <concept>
-<concept_id>10011007.10011006.10011008</concept_id>
-<concept_desc>Software and its engineering~General programming languages</concept_desc>
+<concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
+<concept_desc>Software and its engineering~Functional languages</concept_desc>
 <concept_significance>500</concept_significance>
@@
 <concept>
-<concept_id>10003456.10003457.10003521.10003525</concept_id>
-<concept_desc>Social and professional topics~History of programming languages</concept_desc>
+<concept_id>10011007.10011006.10011008.10011024.10011028</concept_id>
+<concept_desc>Software and its engineering~Data types and structures</concept_desc>
 <concept_significance>300</concept_significance>
+</concept>
+<concept>
+<concept_id>10011007.10011006.10011008.10011024.10011033</concept_id>
+<concept_desc>Software and its engineering~Recursion</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+<concept>
+<concept_id>10003752.10003753.10010622</concept_id>
+<concept_desc>Theory of computation~Abstract machines</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+<concept>
+<concept_id>10003752.10003790.10011740</concept_id>
+<concept_desc>Theory of computation~Type theory</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+<concept>
+<concept_id>10003752.10003790.10002990</concept_id>
+<concept_desc>Theory of computation~Logic and verification</concept_desc>
+<concept_significance>100</concept_significance>
 </concept>
@@
 
-\ccsdesc[500]{Software and its engineering~General programming languages}
-\ccsdesc[300]{Social and professional topics~History of programming languages}
+\ccsdesc[500]{Software and its engineering~Functional languages}
+\ccsdesc[300]{Software and its engineering~Data types and structures}
+\ccsdesc[300]{Software and its engineering~Recursion}
+\ccsdesc[300]{Theory of computation~Abstract machines}
+\ccsdesc[300]{Theory of computation~Type theory}
+\ccsdesc[100]{Theory of computation~Logic and verification}
+
 %% End of generated code

### 2018-06-06 Shake.hs
Added example, fix shake to track figures/preamble

@@ -23,21 +23,14 @@ fmt = "fmt"
 thesis_lagda_files :: [String]
 thesis_lagda_files = [ "main" , "tree" , "problem" , "background"]
 
--- lagda files used for the paper
-paper_lagda_files :: [String]
-paper_lagda_files = [ "main" ]
-
-paper_fmt_files :: [String]
-paper_fmt_files = [ "paper" , "intro" , "generic" ]
-
 main :: IO ()
 main = shakeArgs shakeOptions $ do
   want ["thesis"]
 
   phony "distclean" $ do
-    putNormal "Cleaning files in thesis"
-    cmd_ "latexmk -c -cd thesis/main.tex"
-    removeFilesAfter "thesis" ["*.tex", "*.bbl"]
+    -- putNormal "Cleaning files in thesis"
+    -- cmd_ "latexmk -c -cd thesis/main.tex"
+    -- removeFilesAfter "thesis" ["*.tex", "*.bbl"]
 
     putNormal "Cleaning files in paper"
     cmd_ "latexmk -c -cd paper/main.tex"
@@ -49,13 +42,6 @@ main = shakeArgs shakeOptions $ do
   phony "paper" $ do
     need ["paper/main" <.> pdf]
 
-  "//*.tex" %> \out -> do
-    let input = out -<.> lagda
-        dir   = takeDirectory out
-    putNormal $ "Building " ++ out ++ " from " ++ input ++ " in " ++ dir
-    need [input]
-    cmd_ (Cwd dir) "lhs2TeX --agda -o" [takeFileName out] (takeFileName input)
-
   "thesis/main" <.> pdf %> \out -> do
     let files = [ "thesis" </> file <.> tex | file <- thesis_lagda_files]
         bib   = "thesis/main.bib"
@@ -64,11 +50,21 @@ main = shakeArgs shakeOptions $ do
     need (bib : fmt : sty : files)
     cmd_ "latexmk -f -pdf -cd -xelatex thesis/main.tex"
 
+  "paper/main" <.> tex %> \out -> do
+    let input      = out -<.> lagda
+        dir        = takeDirectory out
+        preamble   = "preamble" <.> tex
+        fmtFiles   = [ "paper" </> file <.> fmt | file <- paper_fmt_files]
+          where paper_fmt_files = [ "paper" , "intro" , "generic" ]
+        sty        = "paper/agda.sty"
+    need $ [input , sty , dir </> "preamble" <.> tex ] ++ fmtFiles
[…]

### 2018-06-07 Shake.hs
Fixed shake clean rules

@@ -27,21 +27,16 @@ main :: IO ()
 main = shakeArgs shakeOptions $ do
   want ["thesis"]
 
-  phony "distclean" $ do
-    -- putNormal "Cleaning files in thesis"
-    -- cmd_ "latexmk -c -cd thesis/main.tex"
-    -- removeFilesAfter "thesis" ["*.tex", "*.bbl"]
-
-    putNormal "Cleaning files in paper"
-    cmd_ "latexmk -c -cd paper/main.tex"
-    removeFilesAfter "paper" ["*.tex", "*.bbl"]
-
   phony "thesis" $ do
     need ["thesis/main" <.> pdf]
 
   phony "paper" $ do
     need ["paper/main" <.> pdf]
 
+  phony "distclean" $ do
+    need ["clean-paper" , "clean-thesis"]
+
+  -- thesis rules
   "thesis/main" <.> pdf %> \out -> do
     let files = [ "thesis" </> file <.> tex | file <- thesis_lagda_files]
         bib   = "thesis/main.bib"
@@ -50,6 +45,7 @@ main = shakeArgs shakeOptions $ do
     need (bib : fmt : sty : files)
     cmd_ "latexmk -f -pdf -cd -xelatex thesis/main.tex"
 
+  -- paper rules
   "paper/main" <.> tex %> \out -> do
     let input      = out -<.> lagda
         dir        = takeDirectory out
@@ -68,3 +64,16 @@ main = shakeArgs shakeOptions $ do
         bib        = "paper/main.bib"
     need $ [main , bib] ++ figures
     cmd_ "latexmk -f -pdf -cd paper/main.tex"
+
+  -- cleaning
+  phony "clean-paper" $ do
+    putNormal "Cleaning files in paper"
+    cmd_ "latexmk -C -cd paper/main.tex"
+    removeFilesAfter "paper" ["main.tex", "*.bbl", "main.pdf", "main.ptb"]
+
+  phony "clean-thesis" $ do
+    return ()
+    -- putNormal "Cleaning files in thesis"
+    -- cmd_ "latexmk -c -cd thesis/main.tex"
+    -- removeFilesAfter "thesis" ["*.tex", "*.bbl"]
+

### 2018-06-08 thesis/figures/figure1.tex
Thesis setup (not compiling latex yet)

--- 2018-06-01 paper/figures/figure3.tex
@@
 \begin{tikzpicture}
-\tikzset{
-    ncbar angle/.initial=90,
-    ncbar/.style={
-        to path=(\tikztostart)
-        -- ($(\tikztostart)!#1!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztotarget)$)
-        -- ($(\tikztotarget)!($(\tikztostart)!#1!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztotarget)$)!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztostart)$)
-        -- (\tikztotarget)
-    },
-    ncbar/.default=0.5cm,
-}
-
-\tikzset{square left brace/.style={ncbar=0.15cm}}
-\tikzset{square right brace/.style={ncbar=-0.15cm}}
   \tikzstyle{every node}=[font=\footnotesize]
@@
   \tikzstyle{level 3}=[level distance=10mm, sibling distance=14mm]
-  \tikzstyle{level 3}=[level distance=10mm, sibling distance=14mm]
   \tikzstyle{load}=[->,thick,color=blue]
   \tikzstyle{unload}=[->,shorten <=1pt,thick,color=red]
-	% \draw[step=1cm, very thin, gray] (-6,-6) grid (6,6);
+
   \node (root) {\AI{Add}}
-  child {node (node1) {\AI{Add}} child {node [label=below:\pgftextcircled{5}] (node2) {\AI{Val}\AS{}\AN{3}}} 
+  child {node (node1) {\AI{Add}} child {node (node2) {\AI{Val}\AS{}\AN{3}}} 
                                  child {node (node3) {\AI{Add}} 
-                                   child {node [label=below:\pgftextcircled{4}] (node4) {\AI{Val}\AS{}\AN{7}}}
-                                   child {node [label=below right:\pgftextcircled{3}] (node5) {\AI{Val}\AS{}\AN{1}}}}} 
-  child {node (node6) {\AI{Add}} child {node [label=below:\pgftextcircled{2}] (node7) {\AI{Val}\AS{}\AN{2}}}
-                                 child {node [label=below right:\pgftextcircled{1}] (node8) {\AI{Val}\AS{}\AN{0}}}};
+                                   child {node (node4) {\AI{Val}\AS{}\AN{7}}}
+                                   child {node (node5) {\AI{Val}\AS{}\AN{1}}}}} 
+  child {node (node6) {\AI{Add}} child {node (node7) {\AI{Val}\AS{}\AN{2}}}
+                                 child {node (node8) {\AI{Val}\AS{}\AN{0}}}};
 
@@
   \draw[unload] (node6) to [ bend right=45] (root);
-  \draw[step=0.1cm,gray] ($(node2)-(0.2,0.2)$) grid ($(node1)-(0.05,-0.15)$);
-  \draw[step=0.1cm,gray] (node2) grid (node4);
-  \draw[step=0.1cm,gray] (node4) grid (node3);
-  \draw[step=0.1cm,gray] (node2) grid ($(node1)!0.25!(node3)$);
-  \draw[step=0.1cm,gray] (node2) grid ($(node1)!0.5!(node3)$);
-  \draw[step=0.1cm,gray] (node2) grid ($(node1)!0.75!(node3)$);
-  \draw[step=0.1cm,gray] (node4) grid ($(node5)!0.75!(node3)$);
-  \draw[step=0.1cm,gray] ($(node4)-(0.2,0.5)$) grid ($(node5)!0.51!(node3)$);
-
-  \draw [thick] (-1.5,-5) to [square left brace ] (-1.5,-4);
-  \draw [thick] (3.5,-5) to [square right brace ] (3.5,-4);
-	\node at (-0.75,-4.5) (right1) {\AI{Right}\AS{}\AN{7},};
-	\node at (0.6,-4.5)    (right2) {\AI{Right}\AS{}\AN{3},};
-  \node at  (1.75,-4.5) {\AI{Left}};
-	\node at (-2.5, -4.5) {\AN{1}\quad,};
-  \tikzstyle{every node}=[font=\tiny]
-  \tikzstyle{level 1}=[level distance=5mm, sibling distance=7mm]
-  \node at (2.75,-4.25) (left1) {\AI{Add}} 
-          child {node {\AI{Val}\AS{}\AN{2}}}
-          child {node {\AI{Val}\AS{}\AN{0}}};
-
-    % \draw [red, thick] (1,0) to [square right brace] (1,4);
 \end{tikzpicture}

### 2018-06-08 thesis/figures/figure2.tex
Thesis setup (not compiling latex yet)

--- 2018-06-01 paper/figures/figure3.tex
@@
   \tikzstyle{unload}=[->,shorten <=1pt,thick,color=red]
-	% \draw[step=1cm, very thin, gray] (-6,-6) grid (6,6);
+
   \node (root) {\AI{Add}}

### 2018-06-08 Shake.hs

@@ -19,10 +19,6 @@ tex = "tex"
 fmt :: FilePath
 fmt = "fmt"
 
--- lagda files used to build the thesis
-thesis_lagda_files :: [String]
-thesis_lagda_files = [ "main" , "tree" , "problem" , "background"]
-
 main :: IO ()
 main = shakeArgs shakeOptions $ do
   want ["thesis"]
@@ -37,12 +33,21 @@ main = shakeArgs shakeOptions $ do
     need ["clean-paper" , "clean-thesis"]
 
   -- thesis rules
+  "thesis/*.tex" %> \out -> do
+    let input     = out -<.> lagda
+        dir       = takeDirectory out
+        fmtFile   = out -<.> fmt
+    need [input , fmtFile]
+    cmd_ (Cwd dir) "lhs2TeX --agda -o" [takeFileName out] (takeFileName input)
+
   "thesis/main" <.> pdf %> \out -> do
-    let files = [ "thesis" </> file <.> tex | file <- thesis_lagda_files]
+    let files = [ "thesis" </> file <.> tex | file <-lagda_files]
+                  where lagda_files = [ "main"      , "introduction" 
+                                      , "background", "expression"
+                                      , "generic"]
         bib   = "thesis/main.bib"
-        fmt   = "thesis/thesis.fmt"
         sty   = "thesis/agda.sty"
-    need (bib : fmt : sty : files)
+    need (bib : sty : files)
     cmd_ "latexmk -f -pdf -cd -xelatex thesis/main.tex"
 
   -- paper rules
@@ -72,8 +77,6 @@ main = shakeArgs shakeOptions $ do
     removeFilesAfter "paper" ["main.tex", "*.bbl", "main.pdf", "main.ptb"]
 
   phony "clean-thesis" $ do
-    return ()
-    -- putNormal "Cleaning files in thesis"
-    -- cmd_ "latexmk -c -cd thesis/main.tex"
-    -- removeFilesAfter "thesis" ["*.tex", "*.bbl"]
-
+    putNormal "Cleaning files in paper"
+    cmd_ "latexmk -C -cd thesis/main.tex"
+    removeFilesAfter "thesis" ["*.tex", "*.bbl", "main.pdf", "main.ptb"]

### 2018-06-21 Shake.hs
Added discussion about a finer grained approach

@@ -47,7 +47,9 @@ main = shakeArgs shakeOptions $ do
                                       , "generic"   , "conclusion" ]
         bib   = "thesis/main.bib"
         sty   = "thesis/agda.sty"
-    need (bib : sty : files)
+        figures    = [ "thesis" </> "figures" </> ("figure" ++ show n) <.> "tex" 
+                     | n <- [1..5]]
+    need (bib : sty : files ++ figures)
     cmd_ "latexmk -f -pdf -cd thesis/main.tex"
 
   -- paper rules

### 2018-07-14 presentation/figures/figure3.tex
[presentation] finished motivation and contrib

\begin{tikzpicture}
  % \tikzstyle{every node}=[font=\footnotesize]
  \tikzstyle{level 1}=[level distance=10mm, sibling distance=24mm]
  \tikzstyle{level 2}=[level distance=10mm, sibling distance=14mm]
  \tikzstyle{level 3}=[level distance=10mm, sibling distance=14mm]
  \tikzstyle{load}=[->,thick,color=blue]
  \tikzstyle{unload}=[->,shorten <=1pt,thick,color=red]

  \node (root) {\AI{Add}}
    child {node (node1) [label=below left:\pgftextcircled{1}] {\AI{Val}\AS{}\AN{1}}} 
    child {node (node2) {\AI{Add}} 
           child {node (node3) [label=below left:\pgftextcircled{2}] {\AI{Val}\AS{}\AN{2}}}
           child {node (node4) [label=below left:\pgftextcircled{3}] {\AI{Val}\AS{}\AN{2}}}};

  {\draw[load] (root) to [ bend right=45] (node1);}
  {\draw[load] (node1) to [ loop below ] (node1);}
  {\draw[unload] (node1) to [ bend left=15] (node2);}
  {\draw[load] (node2) to [ bend right=45] (node3);}
  {\draw[load] (node3) to [ loop below] (node3);}
  {\draw[unload] (node3) to [ bend left=15] (node4);}
  {\draw[load] (node4) to [ loop below] (node4);}
  {\draw[unload] (node4) to [ bend right=45] (node2);}
  {\draw[unload] (node2) to [ bend right=45] (root);}
\end{tikzpicture}

### 2018-07-14 Shake.hs

@@ -19,6 +19,10 @@ tex = "tex"
 fmt :: FilePath
 fmt = "fmt"
 
+thesis_lagda_files = [ "main"      , "introduction" 
+                     , "background", "expression"
+                     , "generic"   , "conclusion" ]
+
 main :: IO ()
 main = shakeArgs shakeOptions $ do
   want ["thesis"]
@@ -44,16 +48,13 @@ main = shakeArgs shakeOptions $ do
     cmd_ (Cwd dir) "lhs2TeX --agda -o" [takeFileName out] (takeFileName input)
 
   "thesis/main" <.> pdf %> \out -> do
-    let files = [ "thesis" </> file <.> tex | file <-lagda_files]
-                  where lagda_files = [ "main"      , "introduction" 
-                                      , "background", "expression"
-                                      , "generic"   , "conclusion" ]
+    let files = [ "thesis" </> file <.> tex | file <- thesis_lagda_files]
         bib   = "thesis/main.bib"
         sty   = "thesis/agda.sty"
         figures    = [ "thesis" </> "figures" </> ("figure" ++ show n) <.> "tex" 
                      | n <- [1..5]]
     need (bib : sty : files ++ figures)
-    cmd_ "latexmk -f -pdf -cd thesis/main.tex"
+    cmd_ "latexmk -f -xelatex -pdf -cd thesis/main.tex"
 
   -- paper rules
   "paper/main" <.> tex %> \out -> do
@@ -80,7 +81,9 @@ main = shakeArgs shakeOptions $ do
     putNormal "Building presentation"
     let input  = out -<.> lagda
         dir    = takeDirectory out
-    need [input, "presentation/presentation.fmt"]
+        figures    = [ "presentation" </> "figures" </> ("figure" ++ show n) <.> "tex" 
+                     | n <- [1..3]]
+    need $ [input, "presentation/presentation.fmt"] ++ figures
     cmd_ (Cwd dir) "lhs2TeX --agda -o" [takeFileName out] (takeFileName input)
      
 
@@ -99,7 +102,8 @@ main = shakeArgs shakeOptions $ do
   phony "clean-thesis" $ do
     putNormal "Cleaning files in paper"
     cmd_ "latexmk -C -cd thesis/main.tex"
-    removeFilesAfter "thesis" ["*.tex", "*.bbl", "main.pdf", "main.ptb"]
+    removeFilesAfter "thesis" $ ["*.bbl", "main.pdf", "main.ptb"]
[…]

### 2018-07-19 presentation/figures/figure3.tex
[presentation] finished

@@
   \tikzstyle{level 1}=[level distance=10mm, sibling distance=24mm]
-  \tikzstyle{level 2}=[level distance=10mm, sibling distance=14mm]
+  \tikzstyle{level 2}=[level distance=10mm, sibling distance=24mm]
   \tikzstyle{level 3}=[level distance=10mm, sibling distance=14mm]
@@
   \node (root) {\AI{Add}}
-    child {node (node1) [label=below left:\pgftextcircled{1}] {\AI{Val}\AS{}\AN{1}}} 
+    child {node (node1) [label=below left:\pgftextcircled{3}] {\AI{Val}\AS{}\AN{1}}} 
     child {node (node2) {\AI{Add}} 
            child {node (node3) [label=below left:\pgftextcircled{2}] {\AI{Val}\AS{}\AN{2}}}
-           child {node (node4) [label=below left:\pgftextcircled{3}] {\AI{Val}\AS{}\AN{2}}}};
+           child {node (node4) [label=below left:\pgftextcircled{1}] {\AI{Val}\AS{}\AN{2}}}};
 

### 2018-08-10 Shake.hs
[paper] change name accent to unicode, update balancing of last page

@@ -100,7 +100,7 @@ main = shakeArgs shakeOptions $ do
   phony "clean-paper" $ do
     putNormal "Cleaning files in paper"
     cmd_ "latexmk -C -cd paper/main.tex"
-    removeFilesAfter "paper" ["main.tex", "*.bbl", "main.pdf", "main.ptb"]
+    removeFilesAfter "paper" ["comment.cut", "main.tex", "*.bbl", "main.ptb"]
 
   phony "clean-thesis" $ do
     putNormal "Cleaning files in paper"

### 2024-10-24 tyde18-presentation/figure.tex
[general] add tyde and Chalmers slides

\begin{tikzpicture}
  % \tikzstyle{every node}=[font=\footnotesize]
  \tikzstyle{level 1}=[level distance=10mm, sibling distance=24mm]
  \tikzstyle{level 2}=[level distance=10mm, sibling distance=14mm]
  \tikzstyle{level 3}=[level distance=10mm, sibling distance=14mm]
  \tikzstyle{load}=[->,thick,color=blue]
  \tikzstyle{unload}=[->,shorten <=1pt,thick,color=red]

  \node (root) {\AI{Add}}
    child {node (node1) {\AI{Val}\AS{}\AN{1}}} 
    child {node (node2) {\AI{Add}} 
           child {node (node3) {\AI{Val}\AS{}\AN{2}}}
           child {node (node4) {\AI{Val}\AS{}\AN{2}}}};

  \onslide<2->{\draw[load] (root) to [ bend right=45] (node1);}
  \onslide<3->{\draw[load] (node1) to [ loop below ] (node1);}
  \onslide<4->{\draw[unload] (node1) to [ bend left=15] (node2);}
  \onslide<5->{\draw[load] (node2) to [ bend right=45] (node3);}
  \onslide<6->{\draw[load] (node3) to [ loop below] (node3);}
  \onslide<7->{\draw[unload] (node3) to [ bend left=15] (node4);}
  \onslide<8->{\draw[load] (node4) to [ loop below] (node4);}
  \onslide<9->{\draw[unload] (node4) to [ bend right=45] (node2);}
  \onslide<11->{\draw[unload] (node2) to [ bend right=45] (root);}
\end{tikzpicture}

### 2024-10-24 Shake.hs

@@ -36,6 +36,12 @@ main = shakeArgs shakeOptions $ do
   phony "presentation" $ do
     need ["presentation/main" <.> pdf]
 
+  phony "tyde18-presentation" $ do
+    need ["tyde18-presentation/main" <.> pdf]
+
+  phony "chalmers-presentation" $ do
+    need ["chalmers-presentation/main" <.> pdf]
+
   phony "distclean" $ do
     need ["clean-paper" , "clean-thesis"]
 
@@ -54,7 +60,7 @@ main = shakeArgs shakeOptions $ do
     let files = [ "thesis" </> file <.> tex | file <- thesis_lagda_files]
         bib   = "thesis/main.bib"
         sty   = "thesis/agda.sty"
-        figures    = [ "thesis" </> "figures" </> ("figure" ++ show n) <.> "tex" 
+        figures    = [ "thesis" </> "figures" </> ("figure" ++ show n) <.> tex
                      | n <- [1..5]]
     need (bib : sty : files ++ figures)
     cmd_ "latexmk -f -xelatex -pdf -cd thesis/main.tex"
@@ -88,13 +94,42 @@ main = shakeArgs shakeOptions $ do
                      | n <- [1..6]]
     need $ [input, "presentation/presentation.fmt"] ++ figures
     cmd_ (Cwd dir) "lhs2TeX --agda -o" [takeFileName out] (takeFileName input)
-     
 
   "presentation/main" <.> pdf %> \out -> do
     putNormal "Building presentation"
     let main  = out -<.> tex
     need [main]
     cmd_ "latexmk -xelatex -pdf -cd presentation/main.tex"
+
+  -- tyde presentation
+  "tyde18-presentation/main" <.> tex %> \out -> do
+    putNormal "Building presentation"
+    let input  = out -<.> lagda
+        dir    = takeDirectory out
+        figures    = [ "tyde18-presentation" </> "figure" <.> "tex" ]
+    need $ [input, "tyde18-presentation/presentation.fmt"] ++ figures
+    cmd_ (Cwd dir) "lhs2TeX --agda -o" [takeFileName out] (takeFileName input)
+
+  "tyde18-presentation/main" <.> pdf %> \out -> do
+    putNormal "Building presentation"
+    let main  = out -<.> tex
+    need [main]
+    cmd_ "latexmk -xelatex -pdf -cd tyde18-presentation/main.tex"
+
+  -- chalmers presentation
[…]

### 2024-10-24 build.sh

@@ -1,2 +1,0 @@
-nix-shell -p "haskell.packages.ghc822.ghcWithPackages (pkgs: [pkgs.shake
-pkgs.lhs2tex])" --command "runhaskell Shake.hs $1"
